---
title: ""
subtitle: ""
author: ""
date: ""
metadata:
  title: "Arabidopsis Growth Under Nutrients and Damage"
format:
  beamer:
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "structurebold"
    navigation: horizontal
    aspectratio: 169
    fontsize: 10pt
    title-slide: false
    header-includes:
      - \usepackage{booktabs}
---

```{r}

source('R/load_arabidopsis.R')

library(dplyr)
library(ggplot2)
library(MASS)
library(car)
library(ggplot2)
library(lme4)
library(lmerTest)
library(knitr)
library(glmmTMB)
library(DHARMa)
library(emmeans)

```

# Arabidopsis Growth Under Nutrients and Damage {.center .vcenter}

**STAT 642 — Spring 2026**  
**Team 11**

**Outline**

1. Introduction & Data Description
2. Model
3. Analysis & Results
4. Assumption Checks
5. Conclusion

# 1.1 Introduction & Data Description

**Source:** Our source dataset is called **`Arabidopsis`**, and was generated
from a **DoE experiment on Arabidopsis thaliana plants** performed by **the Center 
for Genomics and Systems Biology at New York University**, and the results 
of which 
were published in **Banta JA, Stevens MHH, Pigliucci M (2010). “A comprehensive test
of the ’limiting resources’ framework applied to plant tolerance to apical
meristem damage.” Oikos, 119(2), 359-369. doi:10.1111/j.16000706.2009.17726.x.**

Banta et al investigated how different resources **(nutrients, water, and light)**
affect the ability of Arabidopsis plants from 
three different regions to tolerate damage to their **apical 
meristem damage (AMD)** in order to test the "limiting resources" framework. The "limiting resources"  framework was an accepted model in the field, describing 
the growth response of plants to AMD. The dataset contains measurements of plant growth and reproduction under various treatment conditions.

**Sample size:** $n = 625$

**Unit of observation:** individual plants

# 1.2 What Is Interesting About These Data?

- Plant response to **apical meristem damage (AMD)** reflects growth, survival, and adaptation
- AMD occurs due to **herbivory and environmental stress**
- Prior research shows **conflicting results**:
  - Some: higher nutrients → higher tolerance  
  - Others: higher nutrients → lower tolerance
- This dataset allows a **controlled test** of how nutrients affect plant response to damage

**Research Questions**

Our research questions in analyzing this dataset are as follows:

1. How do the reproductive responses of Arabidopsis thaliana to apical meristem damage (AMD) vary across different regions, nutrient levels, and transplant sources?

2. Are there significant interactions between region, nutrient levels, and transplant sources in 
determining the plant's response to AMD?

3. How much of the variability in plant response to AMD can be attributed to differences between 
populations and maternal seed families within regions?

# 1.3 Variables
\small
| Variable | Type | Description |
|----------|------|-------------|
| Total Fruits | Response (integer) | The total number of fruits produced by the plant |
| Region | Fixed predictor | The region from which the plant was collected (NL, SP, SW) |
| Nutrients | Fixed predictor | The level of nutrients in the soil (1 prill = low, 8 prills = high) |
| AMD | Fixed predictor | Whether the plant was clipped to simulate apical meristem damage (clipped or unclipped) |
| Status | Fixed predictor | The source of the plant (Normal, Petri.Plate, or Transplant) |
| Rack | Fixed predictor | The light rack on which the plant was grown (1 or 2) |
| Population | Random grouping | The population of the plant, nested within Region (9 populations) |
| Genotype | Random grouping | The maternal seed family, nested within Population (24 families) |

# 1.4 Experimental Design

- **Study organism:** *Arabidopsis thaliana*
- **Sampling structure:** maternal seed families collected from populations in **3 regions** (Northern Spain, Netherlands, Southern Sweden)

- **Hierarchical sampling structure:**
  - Region
  - Population nested within Region (four populations from SP, two from NL, and three from SW)
  - Genotype (maternal seed family) nested within Population

- **Experimental factors of interest:**
  - **Nutrients:** low (1 prill) vs high (8 prills)
  - **AMD:** clipped vs unclipped
  - **Region:** NL, SP, SW

- **Random factors:**
  - **Population:** 9 populations nested within regions
  - **Genotype:** 24 maternal seed families nested within populations

- **Nuisance factors:**
  - **Rack:** 2 light racks
  - **Status**: Normal, Petri Plate, Transplant (reflecting source of plant material)

- **Experimental details:**

  - Seeds germinated in petri dishes, then planted in pots
  - Failed germinations sometimes replaced by seedlings of the same maternal family
  - Clipping applied at bolting stage to simulate apical meristem damage

- **Design Structure:** A 2 x 2 factorial treatment structure (Nutrients × AMD) applied to plants from a nested sampling hierarchy (Genotypes within populations within regions), with additional fixed effects (Status, Rack) as nuisance factors. Genotype and population are modeled as random
effects while Region is treated as a fixed effect.
- **Sample size:** \(n = 625\)
- **Average replication:** about 6.5 plants per seed-family-by-treatment combination

# 1.5 Data Exploration & Results

Before fitting the model, we perform exploratory data analysis to understand the distribution of the response variable and the relationships between predictors and the response.

::: columns

::: column
```{r, fig.width=4.5, fig.height=3.2, out.width='100%'}

# Distribution of total fruits
ggplot(Arabidopsis, aes(x = total.fruits)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits", x = "Total Fruits", y = "Count") +
  theme_minimal()

```
:::

::: column
```{r, fig.width=4.5, fig.height=3.2, out.width='100%'}
# Distribution of total fruits where total.fruits < 10

ggplot(Arabidopsis %>% filter(total.fruits < 10), aes(x = total.fruits)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits (Zoomed In)", x = "Total Fruits", y = "Count") +
  theme_minimal()
```
:::
:::

# 2.1 Model

**Why a Mixed-Effects Model?**

A mixed-effects model is appropriate because we have a combination of fixed and random effects 
with hierarchical nesting.

- Plants are grouped within **genotypes**
- Genotypes are grouped within **populations**
- Populations are grouped within **regions**

We fit a linear mixed-effects model (LMM) with total fruits
(log-transformed) as the response.

$$
\begin{aligned}
Y_{ijklmno} = \mu &+ \alpha_i + \beta_j + \gamma_k + \delta_l + \phi_m \\
&+ (\alpha\beta)_{ij} + (\alpha\phi)_{im} + (\beta\phi)_{jm} + (\alpha\beta\phi)_{ijm} \\
&+ (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\phi\gamma)_{mk} \\
&+ P_{n(i)} + F_{o(ni)} + \varepsilon_{ijklmno}
\end{aligned}
$$

Main effects:

- Region, Nutrients, AMD, Status (nuisance), Rack (nuisance)

Interactions of interest:

- Region x Nutrients x AMD

- Region × AMD

- Nutrients × AMD

- Region × Nutrients

Nuisance interactions:

- AMD x Status

- Region × Status

- Nutrients × Status

# 2.2 Model Setup

**Fixed effects:**

- $\mu$: overall intercept
- $\alpha_i$: effect of Region ($i = 1, 2, 3$; NL, SP, SW)
- $\beta_j$: effect of Nutrients ($j = 1, 2$; low, high)
- $\gamma_k$: effect of Status/Transplant ($k = 1, 2, 3$; Normal, Petri.Plate, Transplant)
- $\delta_l$: effect of Rack ($l = 1, 2$)
- $\phi_m$: effect of AMD ($m = 1, 2$; clipped, unclipped)
- $(\alpha\beta)_{ij}$: Region $\times$ Nutrients interaction
- $(\alpha\phi)_{im}$: Region $\times$ AMD interaction
- $(\beta\phi)_{jm}$: Nutrients $\times$ AMD interaction
- $(\alpha\beta\phi)_{ijm}$: Region $\times$ Nutrients $\times$ AMD interaction
- $(\alpha\gamma)_{ik}$: Region $\times$ Status interaction (nuisance)
- $(\beta\gamma)_{jk}$: Nutrients $\times$ Status interaction (nuisance)
- $(\phi\gamma)_{mk}$: AMD $\times$ Status interaction (nuisance)

**Random effects:**

- $P_{n(i)} \sim N(0, \sigma^2_P)$: Population $n$ nested within Region $i$
- $F_{o(ni)} \sim N(0, \sigma^2_F)$: Maternal Seed Family $o$ nested within Population and Region
- $\varepsilon_{ijklmno} \sim N(0, \sigma^2)$: residual error

# 2.3 Assumptions

As this data will be analyzed as a linear mixed-effects model, the following assumptions must be met:

1. **Linearity** — the relationship between predictors and each (transformed) response is linear
2. **Independence** — observations are independent conditional on the random effects
3. **Normality** — random effects and residual errors are normally distributed
4. **Equal Variance** — constant residual variance (homoscedasticity)
5. **Correct Model Specification** — the model includes all relevant predictors and interactions, and the random effects structure appropriately captures the nesting and variability in the data

The following additional assumptions are made due to the experimental design:

6. **Random Sampling within Strata** — within each region, populations and maternal seed families are assumed to be representative samples from the broader distribution of populations and families in that region, such that the random effects $P_{n(i)}$ and $F_{o(ni)}$ are exchangeable draws from their respective normal distributions
7. **Exogeneity of Random Effects** — the random effects are uncorrelated with the fixed effects


# 2.4 Model Fitting

:::columns
:::column
**Overview**
- Fit full LMM including:
  - Interactions of interest: Region × Nutrients, Region × AMD,
    Nutrients × AMD, Region × Nutrients × AMD
  - Nuisance interactions: Region × Status, Nutrients × Status,
    AMD × Status

- Use backward elimination (LRT, \( \alpha = 0.05 \))
  → obtain reduced model

- Model fitted with:
  - `lme4::lmer()`  
  - `lmerTest` (Satterthwaite F-tests)

**Response Transformation**

- Use Box-Cox to assess appropriate transformation  
- Result: $\lambda \approx 0.06 \approx 0$ → log transform 
- Use $\log(\text{Total Fruits} + 1)$ to handle zeros
:::
:::column

```{r, fig.width=5, fig.height=5, out.width='100%', echo=FALSE, message=FALSE, warning=FALSE}
# Initial model for Box-Cox assessment
fit_init <- lm(
  (total.fruits + 1) ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status + rack,
  data = Arabidopsis
)

bc <- boxcox(fit_init, lambda = seq(0, 1, by = 0.01))
optimal_lambda <- bc$x[which.max(bc$y)]
abline(v = optimal_lambda, lty = 2)
```

::: 
:::

# 3.1 Analyse

**Type III ANOVA**

```{r}
#| echo: false

Arabidopsis$nutrient <- factor(Arabidopsis$nutrient)
Arabidopsis$rack <- factor(Arabidopsis$rack)
Arabidopsis$logp1_fruits <- log(Arabidopsis$total.fruits + 1)

fit_full <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

#summary(fit_full)
#anova(fit_full, type = 3)
#cat("**Model fit summary**\n\n")
#cat("- Response: log(Total Fruits + 1)\n")
#cat("- Observations: ", nobs(fit_full), "\n", sep = "")
#cat("- Random effects: Population, Genotype\n")
#cat("- Estimation: REML with Satterthwaite F-tests\n")

anova_tab <- anova(fit_full, type = 3)
knitr::kable(
  round(as.data.frame(anova_tab), 3),
  caption = "Type III ANOVA for Full Model"
)
```


# 3.2 Model Reduction

**Backward elimination of non-significant terms** (LRT, $\alpha = 0.05$)

- Test **Nutrients × Status** → not significant, drop
- Test **AMD × Status** → not significant, drop
- Test **Region × Nutrients × AMD** (3-way) → not significant, drop
- Test **Region × AMD** → not significant, drop
- Test **Nutrients × AMD** → not significant, drop

Region × Status and Region × Nutrients remain in the model.

**Final Reduced Model**

```{r}
#| echo: false

# Backward elimination: drop non-significant interactions via LRT
# Start by testing nutrient:status (highest p-value in full model ANOVA)
fit_no_nut_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing full vs reduced (must use ML, not REML)
fit_full_ml <- update(fit_full, REML = FALSE)
fit_no_nut_status_ml <- update(fit_no_nut_status, REML = FALSE)
cat("=== LRT: drop Nutrients x Status ===\n")
lrt1 <- anova(fit_no_nut_status_ml, fit_full_ml)

# anova(fit_no_nut_status, type = 3)

# nutrient:status is non-significant (p = 0.64), so drop it.
# Next: test amd:status in the updated model
fit_no_amd_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing reduced model with amd:status eliminated
fit_no_amd_status_ml <- update(fit_no_amd_status, REML = FALSE)
#cat("\n=== LRT: drop AMD x Status ===\n")
lrt2 <- anova(fit_no_nut_status_ml, fit_no_amd_status_ml)

# anova(fit_no_amd_status, type = 3)

# amd:status is non-significant (p = 0.62), so drop it.
# Next: test removing the reg:nutrient:amd three-way interaction

fit_no_three_way <- lmer(
  logp1_fruits ~ reg * nutrient +
    nutrient * amd +
    reg * amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing reduced model with reg:nutrient:amd eliminated
fit_no_three_way_ml <- update(fit_no_three_way, REML = FALSE)
# cat("\n=== LRT: drop reg:nutrient:amd ===\n")
lrt3 <- anova(fit_no_three_way_ml, fit_no_amd_status_ml)

# anova(fit_no_three_way, type = 3)

# 3FI is non-significant (p = 0.26), so drop it
# Now test removing reg:amd interaction

fit_no_reg_amd <- lmer(
  logp1_fruits ~ reg * nutrient +
    nutrient * amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing reduced model with reg:amd eliminated
fit_no_reg_amd_ml <- update(fit_no_reg_amd, REML = FALSE)
# cat("\n=== LRT: drop reg:amd ===\n")
lrt4 <- anova(fit_no_three_way_ml, fit_no_reg_amd_ml)

# anova(fit_no_reg_amd, type = 3)

# reg:amd non-significant (p = 0.81), so drop it
# Now test removing nutrient:amd interaction

fit_no_nut_amd <- lmer(
  logp1_fruits ~ reg * nutrient +
    amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing reduced model with nutrient:amd eliminated
fit_no_nut_amd_ml <- update(fit_no_nut_amd, REML = FALSE)
# cat("\n=== LRT: drop nutrient:amd ===\n")
lrt5 <- anova(fit_no_reg_amd_ml, fit_no_nut_amd_ml)

anova(fit_no_nut_amd, type = 3)

# After dropping nutrient:amd, all remaining fixed effects are either
# significant or must be retained to preserve hierarchy. Therefore, we
# stop the backward elimination process here.

fit_reduced <- fit_no_nut_amd
anova_reduced <- anova(fit_reduced, type = 3)
#cat("\n=== Reduced Model Summary ===\n")
#summary(fit_reduced)
#cat("\n=== Reduced Model ANOVA ===\n")
#anova(fit_reduced, type = 3)

# ---- Build clean LRT summary table ----
# One row per nested comparison performed during backward elimination.
# Tests are listed in the order they were performed; each row drops one
# interaction from the model selected in the previous row.
lrts <- list(lrt1, lrt2, lrt3, lrt4, lrt5)
lrt_summary <- data.frame(
  Step = seq_along(lrts),
  Test = c(
    "Drop Nutrients × Status",
    "Drop AMD × Status",
    "Drop Region × Nutrients × AMD",
    "Drop Region × AMD",
    "Drop Nutrients × AMD"
  ),
  Chisq   = sapply(lrts, function(x) x$Chisq[2]),
  Df      = sapply(lrts, function(x) x$Df[2]),
  P_Value = sapply(lrts, function(x) x$`Pr(>Chisq)`[2]),
  Decision = "Drop"
)

lrt_summary$Chisq   <- round(lrt_summary$Chisq, 3)
lrt_summary$P_Value <- signif(lrt_summary$P_Value, 3)

# ---- Build reduced-model ANOVA table ----
anova_tbl <- as.data.frame(anova_reduced)
anova_tbl$Effect <- rownames(anova_tbl)
rownames(anova_tbl) <- NULL

anova_tbl <- anova_tbl %>%
  dplyr::select(Effect, `NumDF`, `DenDF`, `F value`, `Pr(>F)`) %>%
  dplyr::mutate(
    DenDF = round(DenDF, 2),
    `F value` = round(`F value`, 2),
    `Pr(>F)` = signif(`Pr(>F)`, 3)
  )

# ---- Print nicely ----
cat("**Backward elimination (LRT, ML fit)**\n\n")
kable(
  lrt_summary,
  align = "clcccc",
  col.names = c("Step", "Test", "Chisq", "Df", "p-value", "Decision")
)

cat("\n\n**Reduced model: Type III tests**\n\n")
kable(
  anova_tbl,
  align = "lcccc",
  col.names = c("Effect", "df1", "df2", "F", "p-value")
)
```

# 3.3 Model Reduction via AIC / BIC

***Alternative model reduction via AIC / BIC***

- Compare all subsets of interaction terms using AIC/BIC
- 7 candidate interactions: 4 treatment (Reg × Nut, Reg × AMD,
  Nut × AMD, Reg × Nut × AMD) and 3 nuisance (Reg × Status,
  Nut × Status, AMD × Status)
- Hierarchy enforced: if the 3-way is included, all three
  corresponding 2-way interactions must also be included
- All main effects retained; Rack included as additive blocking factor
- 72 candidate models (9 treatment-interaction configurations × 8
  nuisance-interaction configurations); fit with ML
- AIC preferred model includes one additional term (Nutrients x AMD) compared to the LRT-reduced model, but this term is not significant in the LRT framework and does not change conclusions about key effects. We retain the LRT-reduced model for interpretability.

```{r}
#| echo: false
#| cache: true


# Main-effects baseline (all main effects + rack + random effects)
base <- "logp1_fruits ~ reg + nutrient + amd + status + rack + (1 | popu) + (1 | gen)"

# Candidate interactions, split into treatment (hierarchy-constrained)
# and nuisance (freely combinable)
treatment_ints <- list(
  "Reg x Nut"       = "reg:nutrient",
  "Reg x AMD"       = "reg:amd",
  "Nut x AMD"       = "nutrient:amd",
  "Reg x Nut x AMD" = "reg:nutrient:amd"
)
nuisance_ints <- list(
  "Reg x Stat" = "reg:status",
  "Nut x Stat" = "nutrient:status",
  "AMD x Stat" = "amd:status"
)

# Enumerate all combinations, filter to hierarchy-respecting subsets.
# If the 3-way is in, all three corresponding 2-way must also be in.
t_combos <- do.call(expand.grid, replicate(4, c(FALSE, TRUE), simplify = FALSE))
names(t_combos) <- names(treatment_ints)
t_valid <- with(t_combos,
  !`Reg x Nut x AMD` | (`Reg x Nut` & `Reg x AMD` & `Nut x AMD`)
)
t_combos <- t_combos[t_valid, , drop = FALSE]

n_combos <- do.call(expand.grid, replicate(3, c(FALSE, TRUE), simplify = FALSE))
names(n_combos) <- names(nuisance_ints)

# Cross treatment x nuisance configurations (9 x 8 = 72)
models <- list()
for (i in seq_len(nrow(t_combos))) {
  for (j in seq_len(nrow(n_combos))) {
    t_in <- names(treatment_ints)[as.logical(t_combos[i, ])]
    n_in <- names(nuisance_ints)[as.logical(n_combos[j, ])]
    included <- c(t_in, n_in)
    if (length(included) == 0) {
      label <- "Main effects only"
      form <- base
    } else {
      label <- paste(included, collapse = " + ")
      terms <- c(treatment_ints[t_in], nuisance_ints[n_in])
      form <- paste(base, paste(terms, collapse = " + "), sep = " + ")
    }
    models[[label]] <- lmer(as.formula(form), data = Arabidopsis, REML = FALSE)
  }
}

aic_bic <- data.frame(
  Model = names(models),
  AIC   = sapply(models, AIC),
  BIC   = sapply(models, BIC)
)

aic_bic$dAIC <- aic_bic$AIC - min(aic_bic$AIC)
aic_bic$dBIC <- aic_bic$BIC - min(aic_bic$BIC)
aic_bic <- aic_bic[order(aic_bic$AIC), ]

# Show top 15 models by AIC (full table would be 72 rows)
knitr::kable(head(aic_bic, 15), digits = 2, row.names = FALSE,
             caption = "Top 15 of 72 candidate models, ranked by AIC")
```

# 3.4 Alternative GLMM with Negative Binomial Family

::: columns

::: column

- **Total Fruits** is a count variable with many zeros → fit a **negative binomial GLMM** on the original response scale

- Same fixed/random effects as the reduced LMM:
  - Main effects: Region, Nutrients, AMD, Status, Rack
  - Interactions: Region × Nutrients, Region × Status
  - Random effects: Population, Genotype

- Goal:
  - assess whether a count-based model provides a better fit
  - compare conclusions with the Gaussian LMM

**Main Result**

- The negative binomial GLMM gave similar conclusions for key fixed effects  
- We retained the LMM as the primary model for interpretability

:::

::: column

```{r}
#| echo: false
#| warning: false
#| message: false

# Fit the same reduced structure as fit_reduced, but with a negative
# binomial response on the original count scale (no log1p transform).
fit_nb <- glmmTMB(
  total.fruits ~ reg * nutrient + reg * status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis,
  family = nbinom2
)

nb_anova <- as.data.frame(car::Anova(fit_nb, type = 3))
nb_anova$Effect <- rownames(nb_anova)
rownames(nb_anova) <- NULL

nb_tbl <- nb_anova[, c("Effect", "Chisq", "Df", "Pr(>Chisq)")]
names(nb_tbl) <- c("Effect", "Chisq", "Df", "p-value")
nb_tbl$Chisq <- round(nb_tbl$Chisq, 2)
nb_tbl$`p-value` <- ifelse(
  nb_tbl$`p-value` < 0.0001,
  "<0.0001",
  formatC(nb_tbl$`p-value`, format = "f", digits = 4)
)

kable(nb_tbl, align = "lccc", caption = "Type III tests for negative binomial GLMM")
```

:::

:::

# 3.5 Variance Components

Research question 3 asks how much variation in plant response is attributable
to differences between populations and maternal seed families within regions.
We answer it by partitioning the random-effect variance from the reduced LMM
and testing each component via LRT.

::: columns

::: column

```{r}
#| echo: false

vc <- as.data.frame(VarCorr(fit_reduced))
vc <- vc[, c("grp", "vcov")]
vc$pct  <- 100 * vc$vcov / sum(vc$vcov)
vc$vcov <- round(vc$vcov, 3)
vc$pct  <- round(vc$pct, 1)
names(vc) <- c("Component", "Variance", "% of total")
kable(
  vc, row.names = FALSE,
  caption = "Variance components (reduced LMM)"
)
```

:::

::: column

```{r}
#| echo: false

rt <- as.data.frame(lmerTest::ranova(fit_reduced))
rt$Term <- rownames(rt)
rt <- rt[, c("Term", "LRT", "Df", "Pr(>Chisq)")]
rt$LRT <- round(rt$LRT, 2)
rt$`Pr(>Chisq)` <- ifelse(
  rt$`Pr(>Chisq)` < 0.001,
  "<0.001",
  formatC(rt$`Pr(>Chisq)`, format = "f", digits = 3)
)
kable(
  rt, row.names = FALSE,
  col.names = c("Term", "LRT", "Df", "p-value"),
  caption = "LRT for random effects"
)
```

:::

:::

Based on the variance components and LRT results, we attribute approximately 16% of the 
variance in fruit production to differences between populations. The LRT for the 
variance contributed by maternal seed families is not significant, suggesting that this 
level of grouping does not explain a significant portion of the variability in the response.

# 4.1 Assumption Checks

::: columns

::: column

**Residual Plots**

- No strong nonlinear pattern  
- Pattern reflects zero-inflation (not model misspecification)  
- Residuals approximately normal  
- Slight variance increase at higher fitted values  

:::

::: column

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
par(mar = c(2.5, 2.5, 0.5, 0.5))
plot(fit_reduced,
     type = c("p", "smooth"),
     col.line = 1,
     xlab = "Fitted values",
     ylab = "Std. residuals",
     main = "")
```
:::

:::

# 4.2 Variance Checks & Conclusion

- Brown–Forsythe (Levene) test:
  - Some heteroscedasticity across Region and Status
  - Variance ratios < 1.5 → not severe

- Large sample size (n = 625):
  - Test is sensitive to small differences*

**Final Assessment**

- Normality: approximately satisfied  
- Independence: satisfied by design  
- Equal variance: mild violations  

→ Overall, model assumptions are **reasonably satisfied**

# 4.3 Assumption Checks: Q–Q plot

::: columns
::: column

- Q–Q plot shows residuals $≈$ normal  
- Shapiro–Wilk test (p = 0.03) attributed to large sample size (n=625) and zero-inflation, not severe non-normality

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
# Formal normality test on residuals
resids <- resid(fit_reduced, scaled = TRUE)
shapiro_test <- shapiro.test(resids)

shapiro_test
```
:::

::: column

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
# Q-Q plot
lattice::qqmath(fit_reduced, main = "Normal Q-Q plot")
```

:::

:::

# 4.4 Assumption Checks: Variance

- Scale-location plot:
  - Slight increase in variance at moderate fitted values  
  - No severe heteroscedasticity pattern  

```{r, fig.width=5, fig.height=2.8, out.width='80%', fig.align='center', echo=FALSE}
scale_loc_plot = function(m) {
  plot(m, sqrt(abs(resid(.))) ~ fitted(.), 
       type = c("p", "smooth"),
       xlab = "Fitted",
       ylab = "Sqrt(|Residuals|)",
       main = "")
}
scale_loc_plot(fit_reduced)

```

# 4.5 Conclusion

::: columns

::: column

**Levene Test**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(car)

lev_results <- data.frame(
  Factor = c("Region", "Nutrients", "Status", "AMD"),
  p_value = c(
    leveneTest(resid(fit_reduced) ~ Arabidopsis$reg)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$nutrient)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$status)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$amd)[1,3]
  )
)

lev_results$p_value <- signif(lev_results$p_value, 3)
knitr::kable(lev_results, align = "lc")

```

**Final Assessment**

```{r, echo=FALSE}
var_reg <- tapply(resid(fit_reduced), Arabidopsis$reg, var)
var_status <- tapply(resid(fit_reduced), Arabidopsis$status, var)

ratio_reg <- max(var_reg) / min(var_reg)
ratio_status <- max(var_status) / min(var_status)

data.frame(
  Group = c("Region", "Status"),
  Variance_Ratio = round(c(ratio_reg, ratio_status), 2)
) %>% knitr::kable(align = "lc")

```

:::

::: column

**Conclusion**

- Log transformation addresses skewness and zeros 

- Mild heteroscedasticity, but not severe  

- GLMM confirms robustness of results  


→ Model assumptions reasonably satisfied  

:::
:::


# 4.6 Hypothesis Tests — Type III F-tests

```{r}
#| echo: false
anova_reduced <- anova(fit_reduced, type = 3)
tbl <- as.data.frame(anova_reduced)
tbl$Effect <- rownames(tbl)
tbl <- tbl[, c("Effect", "F value", "NumDF", "DenDF", "Pr(>F)")]
tbl$`F value` <- round(tbl$`F value`, 3)
tbl$DenDF    <- round(tbl$DenDF, 2)
tbl$`Pr(>F)` <- ifelse(
  tbl$`Pr(>F)` < 0.001,
  "<0.001",
  formatC(tbl$`Pr(>F)`, format = "f", digits = 3)
)
kable(
  tbl, row.names = FALSE,
  col.names = c("Effect", "F", "df1", "df2", "p-value"),
  align = "lcccc",
  caption = "Type III F-tests for log(Total Fruits + 1)"
)
```

**Significant Effectss**

- Significant: **Nutrients, AMD, Rack, and Status**  
- Significant interactions:  
  - Region × Nutrients  
  - Region × Status
  
# 4.7 Post-hoc Analysis (EMMs)

For significant effects, examine **estimated marginal means** on the
log(Total Fruits + 1) scale.

::: columns

::: column

**Nutrients within Region**

```{r}
#| echo: false
emm_nutrient <- emmeans(fit_reduced, pairwise ~ nutrient | reg)
kable(
  as.data.frame(emm_nutrient$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE
)
```

**AMD (main effect)**

```{r}
#| echo: false
emm_amd <- emmeans(fit_reduced, pairwise ~ amd)
kable(
  as.data.frame(emm_amd$contrasts)[, c("contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE
)
```

:::

::: column

**Status within Region**

```{r}
#| echo: false
emm_status <- emmeans(fit_reduced, pairwise ~ status | reg)
kable(
  as.data.frame(emm_status$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE
)
```

:::

:::

# 5.1 Summary of Findings

- **Nutrients, AMD, Rack, and Status** were all found to have significant impacts on fruit 
production. Plants grown in the presence of high nutrients and in the absence of AMD produced significantly more fruits than those under low nutrients and exposed to AMD.

- **Region alone was not significant**, but its role became important through interactions 
with other predictors (Nutrients and Status).

- Overall, fruit production in *Arabidopsis* was shaped by both **experimental treatments** 
and **hierarchical biological differences** among populations and maternal seed families.

# 5.2 Research Question Conclusions

1. How do the reproductive responses of Arabidopsis thaliana to apical meristem damage (AMD) vary across different regions, nutrient levels, and transplant sources?

    - No significant interactions between AMD and Region, Nutrients, or Status were observed, suggesting the effect of AMD on fruit production was consistent across these factors.

2. Are there significant interactions between region, nutrient levels, and transplant sources in determining the a plant's fruit production?

    - Yes, significant interactions were found between Region × Nutrients and Region × Status, indicating that the effects of nutrient availability and transplant source varied across regions.

3. How much of the variability in plant response to AMD can be attributed to differences between populations and maternal seed families within regions?

# 5.3 Limitations \& Future Work

- Residual diagnostics suggested **some heteroscedasticity across regions and status groups**, although the variance differences were not extreme.

- Because the response variable is a **count with many zeros**, transformation-based Gaussian modeling may not fully capture the data structure.

- While the linear mixed-effects model was interpretable and supported the main conclusions, future work should further investigate **generalized mixed models for count outcomes**.

- Additional analyses could explore **interactions involving AMD**, assess alternative variance structures, and use expanded data to better understand variation across populations and genotypes.

# 5.4 References

- Banta, J. A., Stevens, M. H. H., & Pigliucci, M. (2010).  
  *A comprehensive test of the “limiting resources” framework applied to plant tolerance to apical meristem damage*.  
  Oikos, 119(2), 359–369.


- Dean, A., Voss, D., & Draguljić, D. (2017).  
  *Design and Analysis of Experiments* (2nd ed.). Springer.


- STAT 642 Course Handouts, including:
  - Handout 6: Random Effects Models
  - Handout 9: Mixed Models and Variance Components
  - Handout 10: Model Construction and Inference
