---
title: "YOUR TITLE HERE"
subtitle: "STAT 642 — Spring 2026"
author: "Team 11"
date: "April 20, 2026"
format:
  beamer:
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "structurebold"
    navigation: horizontal
    aspectratio: 169
    fontsize: 11pt
    header-includes:
      - \usepackage{booktabs}
---

```{r}

source('R/load_arabidopsis.R')

library(dplyr)
library(ggplot2)

```

## Outline

1. Introduction & Data Description
2. Research Questions
3. Model Specification & Assumptions
4. Analysis & Results
5. Assumption Checks
6. Conclusion

# Introduction & Data Description

## Data Source

- **Source:** Our source dataset is called `Arabadopsis`, and was generated
from a DoE experiment on Arabadopsis thaliana plants performed by the Center 
for Genomics and Systems Biology at New York University, and the results 
of which 
were published in Banta JA, Stevens MHH, Pigliucci M (2010). “A comprehensive test
of the ’limiting resources’ framework applied to plant tolerance to apical
meristem damage.” Oikos, 119(2), 359-369. doi:10.1111/j.16000706.2009.17726.x.

Banta et al investigated how different resources (nutrients, water, and light)
affect the ability of Arabadopsis plants from 
three different regions to tolerate damage to their apical 
meristem (AMD) in order to test the "limiting resources" framework. The "limiting resources"  framework was an accepted model in the field, describing 
the growth response of plants to AMD. The dataset contains measurements of plant growth and reproduction under various treatment conditions.

- **Sample size:** $n = 625$

- **Unit of observation:** individual plants

## Variables

| Variable | Type | Description |
|----------|------|-------------|
| Total Fruits | Response (integer) | The total number of fruits produced by the plant |
| Region | Fixed predictor | The region from which the plant was collected (NL, SP, SW) |
| Nutrients | Fixed predictor | The level of nutrients in the soil (1 prill = low, 8 prills = high) |
| AMD | Fixed predictor | Whether the plant was clipped to simulate apical meristem damage (clipped or unclipped) |
| Status | Fixed predictor | The source of the plant (Normal, Petri.Plate, or Transplant) |
| Rack | Fixed predictor | The light rack on which the plant was grown (1 or 2) |
| Population | Random grouping | The population of the plant, nested within Region (9 populations) |
| Genotype | Random grouping | The maternal seed family, nested within Population (24 families) |

## Experimental Design

Maternal seed families of Arabidopsis thaliana were collected
from populations in three different regions along a broad
latitudinal gradient in western Europe during the spring of
2001. In the laboratory, seeds were hydrated and prepared for germination in petri dishes. 
Seeds were then planted in pots on one of two different high intensity light racks. In cases
where seeds failed to germinate, they were replaced by a transplanted seedling of the same 
maternal family sourced either from a pot with an extra germination or a seed germinated in 
a petri dish. The light exposure and temperature schedule was matched to the natural 
photoperiod and temperature gradient of the region from which the seeds were collected.

Nutrients were added in the form of 14-14-14 nitrogen-phosphorus-potassium (N-P-K) fertilizer
prills. All plants received one nutrient prill 11 days after planting, while plants in the 
high nutrient group recieved an additinal 7 prills 23 days after planting. It was noted that 
there was some variability in the size of the fertilizer prills, which likely contributed 
to residual variance in the growth response of the plants.

At the time of plant bolting, plants were either clipped to simulate apical meristem damage
(AMD) or left unclipped as a control. Clipped plants were all clipped at the same position
with respect to the anatomy of the individual plant. Plants were then analyzed for their
growth response after their reproductive period.

A total of 625 plants were analyzed in this experiment, with an average of 
6.5 plants for each seed family + treatment combination.

- The experiment was a factorial design with the following factors:
  - Region (3 levels: Northern Spain, Netherlands, Southern Sweden)
  - Population (nested within Region, four populations from NS, two from NL, and three from SS)
  - Maternal Seed Family (nested within Population and Region)
  - Nutrients (2 levels: low, high)
  - AMD (2 levels: clipped, unclipped)
  - Rack (2 levels, fixed effect to account for light environment)

# Research Questions

## What Is Interesting About These Data?

The response of plants to apical meristem damage (AMD) is of great interest to plant biologists, 
as it can provide insights into the mechanisms of plant growth and development, as well as the 
evolutionary strategies plants use to cope with damage. Apical meristem damage occurs primarily due 
to plant eating animals, but can also occur due to environmental factors such as wind, hail, or 
freezing temperatures. Understanding how plants respond to AMD can help the scientifific community 
better understand plant resilience and adaptation to environmental stressors, and for this reason, 
the topic is of great interest to plant biologists and ecologists. However, existing research on plant
responses to AMD was somewhat conflicting, with one group of studies reporting that tolerance to AMD
increased with nutrient availability, while another group of studies reported the exact opposite. This 
dataset represents a unique opportunity to systematically investigate the factors that influence plant 
response to AMD and provide more definitive conclusions on the relationship between nutrient 
availability and plant tolerance to AMD.

## Research Questions

Our research questions in analyzing this dataset are as follows:

1. How do the growth and reproductive responses of Arabidopsis thaliana to apical meristem damage (AMD) 
vary across different regions, nutrient levels, and transplant sources?

2. Are there significant interactions between region, nutrient levels, and transplant sources in 
determining the plant's response to AMD?

3. How much of the variability in plant response to AMD can be attributed to differences between 
populations and maternal seed families within regions?

# Model Specification

## Model

We fit a linear mixed-effects model (LMM) with total fruits
(sqrt-transformed) as the response.

$$
Y_{ijklmno} = \mu + \alpha_i + \beta_j + \gamma_k + \delta_l + \phi_m + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + P_{n(i)} + F_{o(ni)} + \varepsilon_{ijklmno}
$$

## Model — Parameter Definitions

**Fixed effects:**

- $\mu$: overall intercept
- $\alpha_i$: effect of Region ($i = 1, 2, 3$; NL, SP, SW)
- $\beta_j$: effect of Nutrients ($j = 1, 2$; low, high)
- $\gamma_k$: effect of Status/Transplant ($k = 1, 2, 3$; Normal, Petri.Plate, Transplant)
- $\delta_l$: effect of Rack ($l = 1, 2$)
- $\phi_m$: effect of AMD ($m = 1, 2$; clipped, unclipped)
- $(\alpha\beta)_{ij}$: Region $\times$ Nutrients interaction
- $(\alpha\gamma)_{ik}$: Region $\times$ Status interaction
- $(\beta\gamma)_{jk}$: Nutrients $\times$ Status interaction

## Model — Random Effects

**Random effects:**

- $P_{n(i)} \sim N(0, \sigma^2_P)$: Population $n$ nested within Region $i$
- $F_{o(ni)} \sim N(0, \sigma^2_F)$: Maternal Seed Family $o$ nested within Population and Region
- $\varepsilon_{ijklmno} \sim N(0, \sigma^2)$: residual error

## Assumptions

As this data will be analyzed as a linear mixed-effects model, the following assumptions must be met:

1. **Linearity** — the relationship between predictors and each (transformed) response is linear
2. **Independence** — observations are independent conditional on the random effects
3. **Normality** — random effects and residual errors are normally distributed
4. **Equal Variance** — constant residual variance (homoscedasticity)
5. **Correct Model Specification** — the model includes all relevant predictors and interactions, and the random effects structure appropriately captures the nesting and variability in the data

The following additional assumptions are made due to the experimental design:

6. **Random Sampling within Strata** — within each region, populations and maternal seed families are assumed to be representative samples from the broader distribution of populations and families in that region, such that the random effects $P_{n(i)}$ and $F_{o(ni)}$ are exchangeable draws from their respective normal distributions
7. **Exogeneity of Random Effects** — the random effects are uncorrelated with the fixed effects

# Analysis & Results

## Data Exploration

Before fitting the model, we perform exploratory data analysis to understand the distribution of the response variable and the relationships between predictors and the response.

```{r}

# Distribution of total fruits
ggplot(Arabidopsis, aes(x = total.fruits)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits", x = "Total Fruits", y = "Count") +
  theme_minimal()

# Distribution of total fruits where total.fruits < 10

ggplot(Arabidopsis %>% filter(total.fruits < 10), aes(x = total.fruits)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits (Zoomed In)", x = "Total Fruits", y = "Count") +
  theme_minimal()

```

## Model Fitting — Overview

We fit the full LMM with all two-way interactions among Region, Nutrients,
and Status, then perform backward elimination of non-significant
interactions using likelihood ratio tests ($\alpha = 0.05$) to arrive at a
reduced model.

We use `lme4::lmer()` for model fitting and `lmerTest` for Satterthwaite
denominator degrees of freedom in F-tests.

## Full Model Fitting

### Box-Cox for full model

As a first step for model fitting, we check the distribution of the response 
variable and consider which transformation might be appropriate in 
order to satisfy model assumptions (namely normality and homoscedasticity). 
We apply the Box-Cox transformation to the response variable (total fruits) and 
assess the suggested lambda values to determine an appropriate transformation. 

```{r}

library(MASS)

# An offset of 1 is added to total fruits to handle zero counts
fit_init <- lm(
  (total.fruits + 1) ~ reg * nutrient * status + amd + rack,
  data = Arabidopsis
)

bc <- boxcox(fit_init, lambda = seq(0, 1, by = 0.01))

optimal_lambda <- bc$x[which.max(bc$y)]
optimal_lambda

```

The Box-Cox transformation reports a lambda value of 0.07, which is close to 0, suggesting that a log transformation may be appropriate. However, since the response variable contains zero counts, we cannot apply a log transformation directly (as log(0) is undefined). Instead, we apply an offset of 1 to each of
the total fruits counts and then apply a log transformation.

```{r}
#| echo: false

library(lme4)
library(lmerTest)

Arabidopsis$nutrient <- factor(Arabidopsis$nutrient)
Arabidopsis$rack <- factor(Arabidopsis$rack)
# Arabidopsis$sqrt_fruits <- sqrt(Arabidopsis$total.fruits)
Arabidopsis$logp1_fruits <- log(Arabidopsis$total.fruits + 1)

fit_full <- lmer(
  logp1_fruits ~ reg * nutrient + reg * status + nutrient * status +
    amd + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
summary(fit_full)
anova(fit_full, type = 3)
```

## Model Reduction

### Backward elimination of non-significant terms

```{r}
#| echo: false

# Backward elimination: drop non-significant interactions via LRT
# Start by testing nutrient:status (highest p-value in full model ANOVA)
fit_no_nut_status <- lmer(
  logp1_fruits ~ reg * nutrient + reg * status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# LRT comparing full vs reduced (must use ML, not REML)
fit_full_ml <- update(fit_full, REML = FALSE)
fit_no_nut_status_ml <- update(fit_no_nut_status, REML = FALSE)
cat("=== LRT: drop Nutrients x Status ===\n")
anova(fit_no_nut_status_ml, fit_full_ml)

# nutrient:status is non-significant (p = 0.64), so drop it
# Now test reg:status in the reduced model
fit_no_reg_status <- lmer(
  logp1_fruits ~ reg * nutrient + status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_reg_status_ml <- update(fit_no_reg_status, REML = FALSE)
cat("\n=== LRT: drop Region x Status ===\n")
anova(fit_no_reg_status_ml, fit_no_nut_status_ml)

# Use the reduced model (without nutrient:status) refit with REML
fit_reduced <- fit_no_nut_status
cat("\n=== Reduced Model Summary ===\n")
summary(fit_reduced)
cat("\n=== Reduced Model ANOVA ===\n")
anova(fit_reduced, type = 3)
```

### Alternative model reduction via AIC / BIC

As an alternative to backward elimination via LRT, we compare all
possible subsets of the two-way interaction terms using AIC and BIC.
With three interactions (Region $\times$ Nutrients, Region $\times$
Status, Nutrients $\times$ Status), this yields $2^3 = 8$ candidate
models. All main effects and covariates are retained in every model.
Models must be fit with ML (not REML) when comparing fixed-effect
structures.

```{r}
#| echo: false

library(knitr)

base <- "logp1_fruits ~ reg + nutrient + status + amd + rack + (1 | popu) + (1 | gen)"

interactions <- list(
  "Reg x Nut"  = "reg:nutrient",
  "Reg x Stat" = "reg:status",
  "Nut x Stat" = "nutrient:status"
)

# Generate all 8 combinations of interactions
combos <- do.call(expand.grid, replicate(3, c(FALSE, TRUE), simplify = FALSE))
names(combos) <- names(interactions)

models <- list()
for (i in seq_len(nrow(combos))) {
  included <- names(interactions)[as.logical(combos[i, ])]
  if (length(included) == 0) {
    label <- "Main effects only"
    form <- base
  } else {
    label <- paste(included, collapse = " + ")
    form <- paste(base, paste(interactions[included], collapse = " + "), sep = " + ")
  }
  models[[label]] <- lmer(as.formula(form), data = Arabidopsis, REML = FALSE)
}

aic_bic <- data.frame(
  Model = names(models),
  AIC   = sapply(models, AIC),
  BIC   = sapply(models, BIC)
)

aic_bic$dAIC <- aic_bic$AIC - min(aic_bic$AIC)
aic_bic$dBIC <- aic_bic$BIC - min(aic_bic$BIC)
aic_bic <- aic_bic[order(aic_bic$AIC), ]

knitr::kable(aic_bic, digits = 2, row.names = FALSE)
```

## Alternative GLMM with Negative Binomial Family

Given that the response is a count variable with many zeros and the heteroscedasticity 
across some variables, we also fit a generalized linear mixed model (GLMM) with a negative 
binomial family to see if it provides a better fit to the data.

```{r}
#| echo: false
#| warning: false
#| message: false

library(glmmTMB)
library(DHARMa)

# Fit the same reduced structure as fit_reduced, but with a negative
# binomial response on the original count scale (no log1p transform).
fit_nb <- glmmTMB(
  total.fruits ~ reg * nutrient + reg * status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis,
  family = nbinom2
)

cat("=== Negative Binomial GLMM Summary ===\n")
summary(fit_nb)

cat("\n=== Type III Wald Chi-square tests ===\n")
car::Anova(fit_nb, type = 3)

# Compare Gaussian LMM on log1p scale vs NB GLMM on count scale via AIC.
# Note: AICs are only directly comparable because glmmTMB reports AIC on
# the original response scale. Use with caution.
cat("\n=== AIC comparison ===\n")
data.frame(
  Model = c("Gaussian LMM (log1p)", "Negative Binomial GLMM"),
  AIC   = c(AIC(fit_reduced), AIC(fit_nb))
)


jacobian_correction <- 2 * sum(log(Arabidopsis$total.fruits + 1))
aic_lmm_adjusted <- AIC(fit_reduced) + jacobian_correction
aic_lmm_adjusted
AIC(fit_nb)

plot(simulateResiduals(fit_nb))
```



# Assumption Checks

## Residual Plots

```{r}
#| echo: false

par(mfrow = c(2, 2))
# Fitted vs Residuals diagnostic plot
plot(fit_reduced, 
     type = c("p", "smooth"), col.line = 1, 
     xlab = "Fitted values", ylab = "Standardized residuals", 
     main = "Fitted vs Residual Plot")
```

The residual vs fitted plot shows an unusual pattern of a diagonal 
line with a slope of -1 due to the presence of zero counts in the response variable, which creates a cluster of residuals at -1 when the fitted values are close to zero. This pattern is a common artifact when modeling count data with a log transformation and zero counts, and it does not necessarily indicate a violation of model assumptions. However, we will further assess the normality and homoscedasticity of the residuals with additional diagnostic plots.

```{r}
# Q-Q plot
lattice::qqmath(fit_reduced, main = "Normal Q-Q plot")

# Formal normality test on residuals
resids <- resid(fit_reduced, scaled = TRUE)
shapiro_test <- shapiro.test(resids)

shapiro_test
```

A linear pattern of residuals in the Q-Q plot indicates that the residuals are 
approximately normally distributed, which supports the normality assumption of the model. 
The Shapiro-Wilk test also fails to reject the null hypothesis of normality (p > 0.05), 
providing further evidence that the residuals do not significantly deviate from a normal 
distribution.

```{r}
library(car)

# Scale-Location plot
# Function found in ?plot.merMod
scale_loc_plot = function(m, line.col = "red", line.lty = 1, line.lwd = 2) {
  plot(fit_reduced, sqrt(abs(resid(.))) ~ fitted(.), 
       type = c("p", "smooth"), 
       xlab = "Fitted values",
       ylab = "Sqrt(Standardized residuals)",
       main = "Scale-Location Plot",
       par.settings = list(plot.line = 
                             list(alpha = 1, 
                                  col = line.col, 
                                  lty = line.lty, 
                                  lwd = line.lwd)))
}
scale_loc_plot(fit_reduced)
```

```{r}

library(car)

# Test constant variance of residuals across levels of each factor
leveneTest(resid(fit_reduced) ~ Arabidopsis$reg)
leveneTest(resid(fit_reduced) ~ Arabidopsis$nutrient)
leveneTest(resid(fit_reduced) ~ Arabidopsis$status)
leveneTest(resid(fit_reduced) ~ Arabidopsis$amd)

```

The Brown-Forsythe-Levene test flags significant heteroscedasticity across the levels of 
Region (p = 0.006) and status (p = 0.025), but not for Nutrients or Status. This suggests 
that the variance of the residuals may differ across regions and culture status, which could 
indicate a violation of the homoscedasticity assumption. 

```{r}

tapply(resid(fit_reduced), Arabidopsis$reg, var)
tapply(resid(fit_reduced), Arabidopsis$status, var)

```

Given the large sample size (n = 625), the BFL test becomes sensitive to relatively small differences in variance across groups, which may not be practically significant.

Reviewing the variance estimates more closely across these two groups, we see that the 
ratio of largest to smallest variance across regions is `r max(tapply(resid(fit_reduced), Arabidopsis$reg, var)) / min(tapply(resid(fit_reduced), Arabidopsis$reg, var))`, and the ratio of largest to smallest variance across status groups is `r max(tapply(resid(fit_reduced), Arabidopsis$status, var)) / min(tapply(resid(fit_reduced), Arabidopsis$status, var))`. While there is some evidence of heteroscedasticity across these groups, the magnitude of the variance differences is not extreme (less than 3-fold difference in variance across groups, 
a rule of thumb cited in some texts), so we will proceed with the analysis while noting this potential violation.

```{r}
#Leverage against standardized residuals
plot(fit_reduced, rstudent(.) ~ hatvalues(.), 
     xlab = "Leverage", 
     ylab = "Standardized Residuals", 
     main = "Leverage vs Residuals")
points()
```

```{r}
#Should be no heteroscedasticity for different levels of random effects
#Check population (popu) and genotype (gen)

#Fitted vs Residuals for population groups
plot(fit_reduced, 
     resid(., scaled = TRUE) ~ fitted(.) | popu,
     abline = 0, 
     xlab = "Fitted values", 
     ylab = "Standardized residuals", 
     main = "Fitted vs Residuals (by Population)")

#Fitted vs Residuals for different levels of genotype
plot(fit_reduced, 
     resid(., scaled = TRUE) ~ fitted(.) | gen,
     abline = 0, 
     xlab = "Fitted values", 
     ylab = "Standardized residuals", 
     main = "Fitted vs Residuals (by Genotype)")
```

```{r}

#Boxplot by popu (population)
plot(fit_reduced, popu ~ resid(., scaled = TRUE), 
     xlab = "Standardized residuals", 
     ylab = "Population")

#Boxplot by gen (genotype)
plot(fit_reduced, gen ~ resid(., scaled = TRUE), 
     xlab = "Standardized residuals", 
     ylab = "Genotype")
```

```{r}
lattice::dotplot(ranef(fit_reduced, condVar = TRUE))$popu

lattice::dotplot(ranef(fit_reduced, condVar = TRUE))$gen
```

## Addressing Violations

<!-- If assumptions are violated, explain remedial measures -->

- TODO — describe any transformations, alternative methods, etc.


# Hypothesis Tests — Type III F-tests

We test the significance of fixed effects using Type III F-tests
with Satterthwaite degrees of freedom.

| Effect | $F$ | df$_1$ | df$_2$ | $p$-value |
|--------|-----|--------|--------|-----------|
| Region | 0.3642 | 2 | 8.73 | 0.705 |
| Nutrients | 149.8963 | 1 | 601.75 | <0.001 |
| Status | 2.0602 | 2 | 606.74 | 0.128 |
| AMD | 6.9777 | 1 | 598.06 | 0.008 |
| Rack | 71.2301 | 1 | 600.21 | <0.001 |
| Region $\times$ Nutrients | 8.7988 | 2 | 600.77 | <0.001 |
| Region $\times$ Status | 2.5587 | 4 | 605.91 | 0.038 |

: Type III F-test results for $\log(\text{Total Fruits} + 1)$

## Significant Differences

When the null hypothesis is rejected for a fixed effect, we examine
estimated marginal means (EMMs) to determine which groups differ and
the direction of the effect.

```{r}
#| echo: false

library(emmeans)
emmeans(fit_reduced, pairwise ~ nutrient | reg)
emmeans(fit_reduced, pairwise ~ status | reg)
emmeans(fit_reduced, pairwise ~ amd)
```

# Conclusion

## Summary of Findings

- TODO — summarize the key results
- TODO — answer the research questions posed earlier

## Limitations & Future Work

- TODO — note any limitations of the analysis
- TODO — suggest directions for future investigation

## References

- TODO — cite data sources and methods as needed
