---
title: ""
subtitle: ""
author: ""
date: ""
metadata:
  title: "Arabidopsis Fruit Production Response to Nutrients and Damage"
format:
  beamer:
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "structurebold"
    navigation: horizontal
    aspectratio: 169
    fontsize: 10pt
    title-slide: false
    header-includes:
      - \usepackage{booktabs}
---

```{r}

library(dplyr)
library(ggplot2)
library(MASS)
library(car)
library(ggplot2)
library(lme4)
library(lmerTest)
library(knitr)
library(glmmTMB)
library(DHARMa)
library(emmeans)

data("Arabidopsis", package = "lme4")

# Sum-to-zero contrasts so Type III ANOVA main-effect tests are
# averaged over factor levels rather than evaluated at a reference level.
options(contrasts = c("contr.sum", "contr.poly"))

# Coerce categorical predictors to factors up front so every model
# (Box-Cox lm, full LMM, NB GLMM) uses the same parameterization.
Arabidopsis$nutrient     <- factor(Arabidopsis$nutrient)
Arabidopsis$rack         <- factor(Arabidopsis$rack)
Arabidopsis$logp1_fruits <- log(Arabidopsis$total.fruits + 1)

```

# Arabidopsis Fruit Production Response to Nutrients and Damage {.center .vcenter}

**STAT 642 — Spring 2026**
**Team 11**

**Outline**

1. Introduction, Data, and Research Questions
2. Model and Assumptions
3. Model Reduction, Diagnostics, and Results
4. Conclusions

# 1. Introduction & Data

::: columns

::: {.column width="55%"}

**Source:** `lme4::Arabidopsis` — a DoE experiment on *Arabidopsis thaliana*
by the Center for Genomics and Systems Biology at NYU, published in
Banta, Stevens, & Pigliucci (2010), *Oikos* 119(2): 359–369.

Banta et al. investigated how **nutrients** and **apical meristem damage (AMD)**
interact to affect fruit production in plants sampled from three European regions,
as a test of the "limiting resources" framework.

- **Sample size:** $n = 625$ plants
- **Response:** Total Fruits (zero-heavy count)
- **Unit of observation:** individual plant

:::

::: {.column width="45%"}

```{r, fig.width=4.2, fig.height=3.0, out.width='100%', echo=FALSE}
ggplot(Arabidopsis, aes(x = total.fruits)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits",
       x = "Total Fruits", y = "Count") +
  theme_minimal()
```

:::

:::

# 2. Research Questions

**Why it matters**

- Plant response to AMD (herbivory / environmental stress) reflects growth, survival, adaptation
- Prior literature is split: higher nutrients → higher *or* lower tolerance
- This dataset enables a controlled test of that trade-off

**Research Questions**

1. How do the reproductive responses of *Arabidopsis thaliana* to AMD vary
   across regions, nutrient levels, and transplant sources?
2. Are there significant interactions between region, nutrient levels, and
   transplant sources in determining fruit production?
3. How much of the variability in fruit production can be attributed to
   differences between populations and maternal seed families within regions?

# 3. Variables & Experimental Design

\footnotesize

```{r}
#| echo: false
vars_tbl <- data.frame(
  Variable = c("Total Fruits", "Region", "Nutrients", "AMD",
               "Status", "Rack", "Population", "Genotype"),
  Role = c("Response", "Fixed", "Fixed", "Fixed",
           "Fixed", "Fixed (additive)", "Random", "Random"),
  Levels = c("count per plant",
             "NL, SP, SW",
             "1 prill (low), 8 prills (high)",
             "clipped, unclipped",
             "Normal, Petri.Plate, Transplant",
             "1, 2",
             "9 levels, nested in Region",
             "24 seed families, nested in Population")
)
kable(vars_tbl, align = "lll", col.names = c("Variable", "Role", "Levels / Description"))
```

\vspace{0.3em}

**Design:** $2 \times 2$ factorial treatment (Nutrients × AMD) crossed with a
3-level sampling hierarchy (Region → Population → Seed family).
Status and Rack are fixed nuisance factors (Rack additive; Status interactions considered).
$\approx$ 6.5 plants per seed-family × treatment cell.

# 4. Model

Fit via `lme4::lmer` with Satterthwaite F-tests (`lmerTest`).
Response: $\log(\text{Total Fruits}+1)$.

$$
\begin{aligned}
Y_{ijklmno} = \mu &+ \alpha_i + \beta_j + \gamma_k + \delta_l + \phi_m \\
&+ (\alpha\beta)_{ij} + (\alpha\phi)_{im} + (\beta\phi)_{jm} + (\alpha\beta\phi)_{ijm} \\
&+ (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\phi\gamma)_{mk} \\
&+ P_{n(i)} + F_{o(ni)} + \varepsilon_{ijklmno}
\end{aligned}
$$

::: columns

::: {.column width="50%"}

\footnotesize

**Fixed effects**

- $\mu$: intercept; $\alpha_i$: Region; $\beta_j$: Nutrients
- $\phi_m$: AMD; $\gamma_k$: Status; $\delta_l$: Rack
- Treatment interactions: $(\alpha\beta), (\alpha\phi), (\beta\phi), (\alpha\beta\phi)$
- Nuisance interactions: $(\alpha\gamma), (\beta\gamma), (\phi\gamma)$

:::

::: {.column width="50%"}

\footnotesize

**Random effects**

- $P_{n(i)} \sim N(0,\sigma^2_P)$: Population in Region
- $F_{o(ni)} \sim N(0,\sigma^2_F)$: Seed family in Population
- $\varepsilon_{ijklmno} \sim N(0,\sigma^2)$: residual

:::

:::

# 5. Assumptions & Response Transformation

::: columns

::: {.column width="55%"}

**LMM assumptions**

1. **Linearity** (in predictors)
2. **Independence** (conditional on random effects)
3. **Normality** of random effects and residuals
4. **Equal variance** of residuals
5. **Correct specification** (fixed + random)
6. **Random sampling within strata** (populations, families)
7. **Exogeneity**: random effects uncorrelated with fixed effects

**Response transformation**

- Box-Cox $\lambda \approx 0.06 \approx 0 \Rightarrow \log$
- Use $\log(\text{Total Fruits}+1)$ to handle zeros

:::

::: {.column width="45%"}

```{r, fig.width=4, fig.height=3.5, out.width='100%', echo=FALSE, message=FALSE, warning=FALSE}
fit_init <- lm(
  (total.fruits + 1) ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status + rack,
  data = Arabidopsis
)

bc <- boxcox(fit_init, lambda = seq(0, 1, by = 0.01))
optimal_lambda <- bc$x[which.max(bc$y)]
abline(v = optimal_lambda, lty = 2)
```

:::

:::

# 6. Model Reduction (LRT)

```{r}
#| echo: false
#| include: false

fit_full <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

# Backward elimination: drop non-significant interactions via LRT
fit_no_nut_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_full_ml <- update(fit_full, REML = FALSE)
fit_no_nut_status_ml <- update(fit_no_nut_status, REML = FALSE)
lrt1 <- anova(fit_no_nut_status_ml, fit_full_ml)

fit_no_amd_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_amd_status_ml <- update(fit_no_amd_status, REML = FALSE)
lrt2 <- anova(fit_no_nut_status_ml, fit_no_amd_status_ml)

fit_no_three_way <- lmer(
  logp1_fruits ~ reg * nutrient + nutrient * amd + reg * amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_three_way_ml <- update(fit_no_three_way, REML = FALSE)
lrt3 <- anova(fit_no_three_way_ml, fit_no_amd_status_ml)

fit_no_reg_amd <- lmer(
  logp1_fruits ~ reg * nutrient + nutrient * amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_reg_amd_ml <- update(fit_no_reg_amd, REML = FALSE)
lrt4 <- anova(fit_no_three_way_ml, fit_no_reg_amd_ml)

fit_no_nut_amd <- lmer(
  logp1_fruits ~ reg * nutrient + amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_nut_amd_ml <- update(fit_no_nut_amd, REML = FALSE)
lrt5 <- anova(fit_no_reg_amd_ml, fit_no_nut_amd_ml)

fit_reduced <- fit_no_nut_amd
anova_reduced <- anova(fit_reduced, type = 3)

# ---- LRT summary ----
lrts <- list(lrt1, lrt2, lrt3, lrt4, lrt5)
lrt_summary <- data.frame(
  Step = seq_along(lrts),
  Test = c(
    "Nutrients × Status",
    "AMD × Status",
    "Region × Nutrients × AMD",
    "Region × AMD",
    "Nutrients × AMD"
  ),
  Chisq   = round(sapply(lrts, function(x) x$Chisq[2]), 2),
  Df      = sapply(lrts, function(x) x$Df[2]),
  P_Value = signif(sapply(lrts, function(x) x$`Pr(>Chisq)`[2]), 3)
)
```

\small

Backward elimination via LRT ($\alpha = 0.05$) on ML-fit models;
drop nuisance interactions first, then highest-order treatment interaction,
then remaining treatment two-ways. **Region × Nutrients** and
**Region × Status** are retained. AIC all-subsets search (72 models,
hierarchy enforced) prefers adding Nutrients × AMD — not LRT-significant, so
dropped for interpretability.

```{r}
#| echo: false
kable(
  lrt_summary,
  align = "clccc",
  col.names = c("Step", "Interaction dropped", "Chisq", "Df", "p"),
  row.names = FALSE
)
```

# 7. Diagnostics

::: columns

::: {.column width="55%"}

```{r, fig.width=4.5, fig.height=1.8, out.width='100%', echo=FALSE}
tight_pad <- list(
  par.xlab.text = list(cex = 0.7),
  par.ylab.text = list(cex = 0.7),
  par.main.text = list(cex = 0.8),
  layout.heights = list(
    top.padding        = 0.2,
    main.key.padding   = 0,
    key.axis.padding   = 0,
    axis.xlab.padding  = 0.2,
    xlab.key.padding   = 0,
    key.sub.padding    = 0,
    bottom.padding     = 0.2
  ),
  layout.widths = list(
    left.padding       = 0.2,
    key.ylab.padding   = 0,
    ylab.axis.padding  = 0.2,
    axis.key.padding   = 0,
    right.padding      = 0.2
  )
)

plot(fit_reduced,
     type = c("p", "smooth"),
     col.line = 1,
     main = list("Residuals vs Fitted", cex = 0.8),
     scales = list(cex = 0.6),
     par.settings = tight_pad)
```

```{r, fig.width=4.5, fig.height=1.8, out.width='100%', echo=FALSE}
lattice::qqmath(fit_reduced,
  main = list("Normal Q-Q", cex = 0.8),
  scales = list(cex = 0.6),
  par.settings = tight_pad)
```

:::

::: {.column width="45%"}

\footnotesize

**Levene test on residuals**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
lev_results <- data.frame(
  Factor = c("Region", "Nutrients", "Status", "AMD"),
  p_value = c(
    leveneTest(resid(fit_reduced) ~ Arabidopsis$reg)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$nutrient)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$status)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$amd)[1,3]
  )
)
lev_results$p_value <- signif(lev_results$p_value, 3)
kable(lev_results, align = "lc")
```

```{r, echo=FALSE}
shapiro_p <- signif(
  shapiro.test(resid(fit_reduced, scaled = TRUE))$p.value, 2
)
```

**Summary**

- Normality: Q–Q approximately linear; Shapiro p $\approx$ `r shapiro_p` reflects large $n$ + zero-inflation
- Variance: mild heteroscedasticity across Region and Status; variance ratios < 1.5
- NB GLMM (slide 12) confirms robustness

:::

:::

# 8. Type III F-tests — Reduced Model

```{r}
#| echo: false
tbl <- as.data.frame(anova_reduced)
tbl$Effect <- rownames(tbl)
tbl <- tbl[, c("Effect", "F value", "NumDF", "DenDF", "Pr(>F)")]
tbl$`F value` <- round(tbl$`F value`, 3)
tbl$DenDF    <- round(tbl$DenDF, 2)
tbl$`Pr(>F)` <- ifelse(
  tbl$`Pr(>F)` < 0.001,
  "<0.001",
  formatC(tbl$`Pr(>F)`, format = "f", digits = 3)
)
kable(
  tbl, row.names = FALSE,
  col.names = c("Effect", "F", "df1", "df2", "p-value"),
  align = "lcccc",
  caption = "Type III F-tests for log(Total Fruits + 1)"
)
```

**Significant effects**

- Main: **Nutrients, AMD, Rack, Status**
- Interactions: **Region × Nutrients**, **Region × Status**

# 9. Variance Components (RQ3)

We partition the random-effect variance from the reduced LMM and test each
component via LRT.

::: columns

::: column

```{r}
#| echo: false

vc <- as.data.frame(VarCorr(fit_reduced))
vc <- vc[, c("grp", "vcov")]
vc$pct  <- 100 * vc$vcov / sum(vc$vcov)
vc$vcov <- round(vc$vcov, 3)
vc$pct  <- round(vc$pct, 1)
names(vc) <- c("Component", "Variance", "% of total")
kable(
  vc, row.names = FALSE,
  caption = "Variance components (reduced LMM)"
)

popu_pct <- vc$`% of total`[vc$Component == "popu"]
```

:::

::: column

```{r}
#| echo: false

rt <- as.data.frame(lmerTest::ranova(fit_reduced))
rt$Term <- rownames(rt)
rt <- rt[, c("Term", "LRT", "Df", "Pr(>Chisq)")]
rt$LRT <- round(rt$LRT, 2)
rt$`Pr(>Chisq)` <- ifelse(
  rt$`Pr(>Chisq)` < 0.001,
  "<0.001",
  formatC(rt$`Pr(>Chisq)`, format = "f", digits = 3)
)
kable(
  rt, row.names = FALSE,
  col.names = c("Term", "LRT", "Df", "p-value"),
  caption = "LRT for random effects"
)
```

:::

:::

Approximately **`r popu_pct`%** of the variance in fruit production is attributable to
differences between populations. The LRT for maternal seed families is not
significant, so that level does not explain a meaningful portion of the variability.

# 10. Post-hoc Analysis (EMMs)

Pairwise contrasts on the $\log(\text{Total Fruits}+1)$ scale.

::: columns

::: {.column width="48%"}

\footnotesize

**Nutrients within Region**

```{r}
#| echo: false
emm_nutrient <- emmeans(fit_reduced, pairwise ~ nutrient | reg)
kable(
  as.data.frame(emm_nutrient$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Reg", "Contrast", "Est.", "SE", "p")
)
```

\vspace{0.6em}

**AMD (main effect)**

```{r}
#| echo: false
emm_amd <- emmeans(fit_reduced, pairwise ~ amd)
kable(
  as.data.frame(emm_amd$contrasts)[, c("contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Contrast", "Est.", "SE", "p")
)
```

:::

::: {.column width="52%"}

\footnotesize

**Status within Region**

```{r}
#| echo: false
emm_status <- emmeans(fit_reduced, pairwise ~ status | reg)
kable(
  as.data.frame(emm_status$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Reg", "Contrast", "Est.", "SE", "p")
)
```

:::

:::

# 11. Interaction Plots

::: columns

::: column

**Region × Nutrients**

```{r, fig.width=4.5, fig.height=3.5, out.width='100%', echo=FALSE}
Arabidopsis %>%
  group_by(reg, nutrient) %>%
  summarise(
    mean_fruits = mean(logp1_fruits),
    se = sd(logp1_fruits) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = factor(nutrient, labels = c("Low (1 prill)", "High (8 prills)")),
             y = mean_fruits,
             color = reg,
             group = reg)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_fruits - se, ymax = mean_fruits + se),
                width = 0.1, linewidth = 0.8) +
  labs(x = "Nutrient Level",
       y = "Mean log(Total Fruits + 1)",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 10))
```

- All regions benefit from high nutrients
- SP: steepest increase; SW: smallest response

:::

::: column

**Region × Status**

```{r, fig.width=4.5, fig.height=3.5, out.width='100%', echo=FALSE}
Arabidopsis %>%
  group_by(reg, status) %>%
  summarise(
    mean_fruits = mean(logp1_fruits),
    se = sd(logp1_fruits) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = status,
             y = mean_fruits,
             color = reg,
             group = reg)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_fruits - se, ymax = mean_fruits + se),
                width = 0.15, linewidth = 0.8) +
  labs(x = "Plant Status",
       y = "Mean log(Total Fruits + 1)",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        axis.text.x = element_text(angle = 15, hjust = 1))
```

- SP: Normal plants outperform Petri.Plate
- SW: Petri.Plate lowest; NL: less differentiation

:::

:::

# 12. Robustness: Negative Binomial GLMM

::: columns

::: {.column width="45%"}

\small

- Total Fruits is a zero-heavy count — fit a **negative binomial GLMM**
  on the original scale
- Same fixed / random effects as the reduced LMM
- **Result:** conclusions for key fixed effects align with the LMM;
  LMM retained for interpretability

:::

::: {.column width="55%"}

\footnotesize

```{r}
#| echo: false
#| warning: false
#| message: false

fit_nb <- glmmTMB(
  total.fruits ~ reg * nutrient + reg * status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis,
  family = nbinom2
)

nb_anova <- as.data.frame(car::Anova(fit_nb, type = 3))
nb_anova$Effect <- rownames(nb_anova)
rownames(nb_anova) <- NULL

nb_tbl <- nb_anova[, c("Effect", "Chisq", "Df", "Pr(>Chisq)")]
names(nb_tbl) <- c("Effect", "Chisq", "Df", "p-value")
nb_tbl$Chisq <- round(nb_tbl$Chisq, 2)
nb_tbl$`p-value` <- ifelse(
  nb_tbl$`p-value` < 0.0001,
  "<0.0001",
  formatC(nb_tbl$`p-value`, format = "f", digits = 4)
)

kable(nb_tbl, align = "lccc",
      caption = "Type III tests: NB GLMM")
```

:::

:::

# 13. Conclusions

**Summary**

- **Nutrients, AMD, Rack, Status** are significant; high nutrients and absence
  of AMD → more fruits
- **Region alone** is not significant as a main effect but matters through
  **Region × Nutrients** and **Region × Status** interactions

**Answers to research questions**

1. *AMD response across region / nutrients / status:* No significant interactions
   between AMD and those factors — AMD's effect is consistent across them.
2. *Other interactions:* **Region × Nutrients** and **Region × Status** are
   significant — nutrient and transplant-source effects vary by region.
3. *Hierarchical variance:* $\approx$ `r popu_pct`% of variance attributed
   to populations; seed-family variance not significant.

# 14. Limitations, Future Work, & References

**Limitations & future work**

- Mild heteroscedasticity across Region / Status (variance ratios < 1.5)
- Zero-heavy count response → Gaussian LMM on $\log(y+1)$ is approximate;
  NB GLMM corroborates conclusions but should be the primary model in future work

\vspace{0.5em}

**References**

\footnotesize

- Banta, J. A., Stevens, M. H. H., & Pigliucci, M. (2010). *A comprehensive test
  of the "limiting resources" framework applied to plant tolerance to apical
  meristem damage.* Oikos, 119(2), 359–369.
- Dean, A., Voss, D., & Draguljić, D. (2017). *Design and Analysis of Experiments*
  (2nd ed.). Springer.
- STAT 642 Course Handouts 6 (Random Effects), 9 (Mixed Models and Variance
  Components), 10 (Model Construction and Inference).

# Backup Slides {.center .vcenter}

Reference material for Q\&A

# B1. Full Model: Type III ANOVA

Before reduction, the full LMM includes all treatment interactions up to 3-way
and all nuisance interactions with Status.

\footnotesize

```{r}
#| echo: false
anova_full <- anova(fit_full, type = 3)
full_tbl <- as.data.frame(anova_full)
full_tbl$Effect <- rownames(full_tbl)
full_tbl <- full_tbl[, c("Effect", "F value", "NumDF", "DenDF", "Pr(>F)")]
full_tbl$`F value` <- round(full_tbl$`F value`, 3)
full_tbl$DenDF    <- round(full_tbl$DenDF, 2)
full_tbl$`Pr(>F)` <- ifelse(
  full_tbl$`Pr(>F)` < 0.001,
  "<0.001",
  formatC(full_tbl$`Pr(>F)`, format = "f", digits = 3)
)
kable(
  full_tbl, row.names = FALSE,
  col.names = c("Effect", "F", "df1", "df2", "p-value"),
  align = "lcccc",
  caption = "Type III F-tests for the full model"
)
```

# B2. Model Reduction via AIC / BIC

```{r}
#| echo: false
#| include: false
#| cache: true

# Main-effects baseline (all main effects + rack + random effects)
base <- "logp1_fruits ~ reg + nutrient + amd + status + rack + (1 | popu) + (1 | gen)"

treatment_ints <- list(
  "Reg x Nut"       = "reg:nutrient",
  "Reg x AMD"       = "reg:amd",
  "Nut x AMD"       = "nutrient:amd",
  "Reg x Nut x AMD" = "reg:nutrient:amd"
)
nuisance_ints <- list(
  "Reg x Stat" = "reg:status",
  "Nut x Stat" = "nutrient:status",
  "AMD x Stat" = "amd:status"
)

# Hierarchy: if 3-way is in, all three 2-way must also be in.
t_combos <- do.call(expand.grid, replicate(4, c(FALSE, TRUE), simplify = FALSE))
names(t_combos) <- names(treatment_ints)
t_valid <- with(t_combos,
  !`Reg x Nut x AMD` | (`Reg x Nut` & `Reg x AMD` & `Nut x AMD`)
)
t_combos <- t_combos[t_valid, , drop = FALSE]

n_combos <- do.call(expand.grid, replicate(3, c(FALSE, TRUE), simplify = FALSE))
names(n_combos) <- names(nuisance_ints)

models <- list()
for (i in seq_len(nrow(t_combos))) {
  for (j in seq_len(nrow(n_combos))) {
    t_in <- names(treatment_ints)[as.logical(t_combos[i, ])]
    n_in <- names(nuisance_ints)[as.logical(n_combos[j, ])]
    included <- c(t_in, n_in)
    if (length(included) == 0) {
      label <- "Main effects only"
      form <- base
    } else {
      label <- paste(included, collapse = " + ")
      terms <- c(treatment_ints[t_in], nuisance_ints[n_in])
      form <- paste(base, paste(terms, collapse = " + "), sep = " + ")
    }
    models[[label]] <- lmer(as.formula(form), data = Arabidopsis, REML = FALSE)
  }
}

aic_bic <- data.frame(
  Model = names(models),
  AIC   = sapply(models, AIC),
  BIC   = sapply(models, BIC)
)
aic_bic$dAIC <- aic_bic$AIC - min(aic_bic$AIC)
aic_bic$dBIC <- aic_bic$BIC - min(aic_bic$BIC)
aic_bic <- aic_bic[order(aic_bic$AIC), ]
```

::: columns

::: {.column width="42%"}

\small

- Enumerate all subsets of 7 candidate interactions
- **Hierarchy enforced**: if 3-way is in, all three 2-way interactions must also be in
- Main effects always retained; Rack additive
- 72 models (9 treatment × 8 nuisance); fit with ML
- AIC-preferred model adds **Nutrients × AMD** vs. the LRT-reduced model;
  not significant by LRT
- Retain LRT-reduced model for interpretability

:::

::: {.column width="58%"}

\footnotesize

```{r}
#| echo: false
kable(
  head(aic_bic, 10), digits = 2, row.names = FALSE,
  caption = "Top 10 of 72 models, ranked by AIC"
)
```

:::

:::

# B3. Scale-Location Plot

- Slight increase in variance at moderate fitted values
- No severe heteroscedasticity pattern

```{r, fig.width=5, fig.height=2.8, out.width='80%', fig.align='center', echo=FALSE}
scale_loc_plot <- function(m) {
  plot(m, sqrt(abs(resid(.))) ~ fitted(.),
       type = c("p", "smooth"),
       xlab = "Fitted",
       ylab = "Sqrt(|Residuals|)",
       main = "")
}
scale_loc_plot(fit_reduced)
```

# B4. Normality Detail: Shapiro-Wilk

::: columns

::: column

\footnotesize

```{r, echo=FALSE}
resids <- resid(fit_reduced, scaled = TRUE)
shapiro_test <- shapiro.test(resids)
shapiro_test
```

- With $n = 625$, Shapiro-Wilk is sensitive to tiny departures from normality.
- Q-Q plot is the primary diagnostic; residuals approximately normal.
- The mild non-normality is attributed to zero-inflation and is addressed
  indirectly by the NB GLMM robustness check.

:::

::: column

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
lattice::qqmath(fit_reduced, main = "Normal Q-Q plot")
```

:::

:::

# B5. Variance Diagnostics Detail

::: columns

::: column

**Levene (Brown-Forsythe) test on residuals**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
lev_results <- data.frame(
  Factor = c("Region", "Nutrients", "Status", "AMD"),
  p_value = c(
    leveneTest(resid(fit_reduced) ~ Arabidopsis$reg)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$nutrient)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$status)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$amd)[1,3]
  )
)
lev_results$p_value <- signif(lev_results$p_value, 3)
kable(lev_results, align = "lc")
```

**Variance ratios**

```{r, echo=FALSE}
var_reg    <- tapply(resid(fit_reduced), Arabidopsis$reg, var)
var_status <- tapply(resid(fit_reduced), Arabidopsis$status, var)
ratio_reg    <- max(var_reg) / min(var_reg)
ratio_status <- max(var_status) / min(var_status)

data.frame(
  Group = c("Region", "Status"),
  Variance_Ratio = round(c(ratio_reg, ratio_status), 2)
) %>% kable(align = "lc")
```

:::

::: column

**Interpretation**

- Levene detects some heteroscedasticity across Region and Status
- Variance ratios < 1.5 → not severe
- Large $n$ makes the test sensitive to small differences
- Log transformation addresses skewness and zeros
- NB GLMM confirms robustness of conclusions

:::

:::

# B6. Response Distribution (Zoomed)

```{r, fig.width=6, fig.height=3.2, out.width='70%', fig.align='center', echo=FALSE}
ggplot(Arabidopsis %>% filter(total.fruits < 10), aes(x = total.fruits)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits (zoomed to < 10)",
       x = "Total Fruits", y = "Count") +
  theme_minimal()
```

- Heavy mass at zero motivates the $\log(y+1)$ transformation and the NB GLMM
  robustness check.

# B7. Sampling Hierarchy Detail

::: columns

::: column

**Hierarchy**

- 3 regions: NL, SP, SW
- 9 populations nested in Region (4 SP, 2 NL, 3 SW)
- 24 maternal seed families nested in Population

**Procedure**

- Seeds germinated in petri dishes, then potted
- Failed germinations replaced with same-family seedlings
- Clipping (AMD) applied at bolting stage

:::

::: column

**Random-effect specification**

- `(1 | popu)` — Population random intercept;
  levels uniquely labeled so nesting in Region is implicit
- `(1 | gen)` — Seed family random intercept;
  levels uniquely labeled so nesting in Population is implicit

**Scale**

- $n = 625$ plants
- $\approx$ 6.5 plants per seed-family $\times$ treatment cell

:::

:::
