---
title: ""
subtitle: ""
author: ""
date: ""
metadata:
  title: "Arabidopsis Fruit Production Under Nutrients and Damage"
format:
  beamer:
    theme: "Madrid"
    colortheme: "default"
    fonttheme: "structurebold"
    navigation: horizontal
    aspectratio: 169
    fontsize: 10pt
    title-slide: false
    header-includes:
      - \usepackage{booktabs}
---

```{r}

source('R/load_arabidopsis.R')

library(dplyr)
library(ggplot2)
library(MASS)
library(car)
library(ggplot2)
library(lme4)
library(lmerTest)
library(knitr)
library(glmmTMB)
library(DHARMa)
library(emmeans)

# Sum-to-zero contrasts so Type III ANOVA main-effect tests are
# averaged over factor levels rather than evaluated at a reference level.
options(contrasts = c("contr.sum", "contr.poly"))

# Coerce categorical predictors to factors up front so every model
# (Box-Cox lm, full LMM, NB GLMM) uses the same parameterization.
Arabidopsis$nutrient     <- factor(Arabidopsis$nutrient)
Arabidopsis$rack         <- factor(Arabidopsis$rack)
Arabidopsis$logp1_fruits <- log(Arabidopsis$total.fruits + 1)

```

# Arabidopsis Growth Under Nutrients and Damage {.center .vcenter}

**STAT 642 — Spring 2026**  
**Team 11**

**Outline**

1. Introduction & Data Description
2. Model
3. Analysis & Results
4. Assumption Checks
5. Conclusion

# 1.1 Introduction & Data Description

**Source:** Our source dataset is called **`Arabidopsis`**, and was generated
from a **DoE experiment on Arabidopsis thaliana plants** performed by **the Center 
for Genomics and Systems Biology at New York University**, and the results 
of which 
were published in **Banta JA, Stevens MHH, Pigliucci M (2010). “A comprehensive test
of the ’limiting resources’ framework applied to plant tolerance to apical
meristem damage.” Oikos, 119(2), 359-369. doi:10.1111/j.16000706.2009.17726.x.**

Banta et al investigated how different resources **(nutrients and light)**
affect the ability of Arabidopsis plants from 
three different regions to tolerate damage to their **apical 
meristem (AMD)** in order to test the "limiting resources" framework. The "limiting resources"  framework was an accepted model in the field, describing 
the growth response of plants to AMD. The dataset contains measurements of plant growth and reproduction under various treatment conditions.

**Sample size:** $n = 625$

**Unit of observation:** individual plants

# 1.2 What Is Interesting About These Data?

- Plant response to **apical meristem damage (AMD)** reflects growth, survival, and adaptation
- AMD occurs due to **herbivory and environmental stress**
- Prior research shows **conflicting results**:
  - Some: higher nutrients → higher tolerance  
  - Others: higher nutrients → lower tolerance
- This dataset allows a **controlled test** of how nutrients affect plant response to damage

**Research Questions**

Our research questions in analyzing this dataset are as follows:

1. How do the reproductive responses of Arabidopsis thaliana to apical meristem damage (AMD) vary across different regions, nutrient levels, and transplant sources?

2. Are there significant interactions between region, nutrient levels, and transplant sources in 
determining the plant's response to AMD?

3. How much of the variability in fruit production can be attributed to differences between 
populations and maternal seed families within regions?

# 1.3 Variables
\small
| Variable | Type | Description |
|----------|------|-------------|
| Total Fruits | Response (integer) | The total number of fruits produced by the plant |
| Region | Fixed predictor | The region from which the plant was collected (NL, SP, SW) |
| Nutrients | Fixed predictor | The level of nutrients in the soil (1 prill = low, 8 prills = high) |
| AMD | Fixed predictor | Whether the plant was clipped to simulate apical meristem damage (clipped or unclipped) |
| Status | Fixed predictor | The source of the plant (Normal, Petri.Plate, or Transplant) |
| Rack | Fixed predictor | The light rack on which the plant was grown (1 or 2) |
| Population | Random grouping | The population of the plant, nested within Region (9 populations) |
| Genotype | Random grouping | The maternal seed family, nested within Population (24 families) |

# 1.4 Experimental Design

::: columns

::: column

**Study organism:** *Arabidopsis thaliana*

**Sampling hierarchy**

- 3 regions: NL, SP, SW
- 9 populations nested in Region (4 SP, 2 NL, 3 SW)
- 24 maternal seed families nested in Population

**Treatment factors**

- Nutrients: 1 prill (low) vs 8 prills (high)
- AMD: clipped vs unclipped
- Region: NL, SP, SW (fixed)

**Nuisance factors (fixed)**

- Rack: 2 light racks (interactions are not considered)
- Status: Normal / Petri.Plate / Transplant (interactions are considered)

:::

::: column

**Random factors**

- Population (9 levels)
- Genotype / maternal seed family (24 levels)

**Design structure**

A 2 × 2 factorial treatment (Nutrients × AMD) crossed with a
3-level sampling hierarchy. Region is fixed; Population and
Genotype are random.

**Procedure**

- Seeds germinated in petri dishes, then potted
- Failed germinations replaced with same-family seedlings
- Clipping applied at bolting stage

**Scale:** \(n = 625\); ≈ 6.5 plants per seed-family × treatment cell

:::

:::

# 1.5 Data Exploration & Results

Before fitting the model, we perform exploratory data analysis to understand the distribution of the response variable and the relationships between predictors and the response.

::: columns

::: column
```{r, fig.width=4.5, fig.height=3.2, out.width='100%'}

# Distribution of total fruits
ggplot(Arabidopsis, aes(x = total.fruits)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits", x = "Total Fruits", y = "Count") +
  theme_minimal()

```
:::

::: column
```{r, fig.width=4.5, fig.height=3.2, out.width='100%'}
# Distribution of total fruits where total.fruits < 10

ggplot(Arabidopsis %>% filter(total.fruits < 10), aes(x = total.fruits)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Total Fruits (Zoomed In)", x = "Total Fruits", y = "Count") +
  theme_minimal()
```
:::
:::

# 2.1 Model

**Why a Mixed-Effects Model?**

Fixed and random effects are combined to reflect the nesting:
plants within **genotypes**, genotypes within **populations**,
populations within **regions**.

Response: $\log(\text{Total Fruits} + 1)$; fit via
`lme4::lmer` with Satterthwaite F-tests (`lmerTest`).

$$
\begin{aligned}
Y_{ijklmno} = \mu &+ \alpha_i + \beta_j + \gamma_k + \delta_l + \phi_m \\
&+ (\alpha\beta)_{ij} + (\alpha\phi)_{im} + (\beta\phi)_{jm} + (\alpha\beta\phi)_{ijm} \\
&+ (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\phi\gamma)_{mk} \\
&+ P_{n(i)} + F_{o(ni)} + \varepsilon_{ijklmno}
\end{aligned}
$$

Term definitions appear on the next slide.

# 2.2 Model Setup

::: columns

::: column

**Fixed main effects**

- $\mu$: overall intercept
- $\alpha_i$: Region ($i=1,2,3$; NL, SP, SW)
- $\beta_j$: Nutrients ($j=1,2$; low, high)
- $\phi_m$: AMD ($m=1,2$; clipped, unclipped)
- $\gamma_k$: Status ($k=1,2,3$)
- $\delta_l$: Rack ($l=1,2$)

**Random effects**

- $P_{n(i)} \sim N(0, \sigma^2_P)$: Population in Region
- $F_{o(ni)} \sim N(0, \sigma^2_F)$: Seed family in Population
- $\varepsilon_{ijklmno} \sim N(0, \sigma^2)$: residual error

:::

::: column

**Treatment interactions**

- $(\alpha\beta)_{ij}$: Region × Nutrients
- $(\alpha\phi)_{im}$: Region × AMD
- $(\beta\phi)_{jm}$: Nutrients × AMD
- $(\alpha\beta\phi)_{ijm}$: Region × Nutrients × AMD

**Nuisance interactions**

- $(\alpha\gamma)_{ik}$: Region × Status
- $(\beta\gamma)_{jk}$: Nutrients × Status
- $(\phi\gamma)_{mk}$: AMD × Status

:::

:::

# 2.3 Assumptions

As this data will be analyzed as a linear mixed-effects model, the following assumptions must be met:

1. **Linearity** — the relationship between predictors and each (transformed) response is linear
2. **Independence** — observations are independent conditional on the random effects
3. **Normality** — random effects and residual errors are normally distributed
4. **Equal Variance** — constant residual variance (homoscedasticity)
5. **Correct Model Specification** — the model includes all relevant predictors and interactions, and the random effects structure appropriately captures the nesting and variability in the data

The following additional assumptions are made due to the experimental design:

6. **Random Sampling within Strata** — within each region, populations and maternal seed families are assumed to be representative samples from the broader distribution of populations and families in that region, such that the random effects $P_{n(i)}$ and $F_{o(ni)}$ are exchangeable draws from their respective normal distributions
7. **Exogeneity of Random Effects** — the random effects are uncorrelated with the fixed effects


# 2.4 Model Fitting

:::columns
:::column
**Overview**
- Fit full LMM including:
  - Interactions of interest: Region × Nutrients, Region × AMD,
    Nutrients × AMD, Region × Nutrients × AMD
  - Nuisance interactions: Region × Status, Nutrients × Status,
    AMD × Status

- Use backward elimination (LRT, \( \alpha = 0.05 \))
  → obtain reduced model

- Model fitted with:
  - `lme4::lmer()`  
  - `lmerTest` (Satterthwaite F-tests)

**Response Transformation**

- Use Box-Cox to assess appropriate transformation  
- Result: $\lambda \approx 0.06 \approx 0$ → log transform 
- Use $\log(\text{Total Fruits} + 1)$ to handle zeros
:::
:::column

```{r, fig.width=5, fig.height=5, out.width='100%', echo=FALSE, message=FALSE, warning=FALSE}
# Initial model for Box-Cox assessment
fit_init <- lm(
  (total.fruits + 1) ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status + rack,
  data = Arabidopsis
)

bc <- boxcox(fit_init, lambda = seq(0, 1, by = 0.01))
optimal_lambda <- bc$x[which.max(bc$y)]
abline(v = optimal_lambda, lty = 2)
```

::: 
:::

# 3.1 Analysis

**Type III ANOVA**

```{r}
#| echo: false

fit_full <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + nutrient * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)

#summary(fit_full)
#anova(fit_full, type = 3)
#cat("**Model fit summary**\n\n")
#cat("- Response: log(Total Fruits + 1)\n")
#cat("- Observations: ", nobs(fit_full), "\n", sep = "")
#cat("- Random effects: Population, Genotype\n")
#cat("- Estimation: REML with Satterthwaite F-tests\n")

anova_tab <- anova(fit_full, type = 3)
knitr::kable(
  round(as.data.frame(anova_tab), 3),
  caption = "Type III ANOVA for Full Model"
)
```


# 3.2 Model Reduction

Backward elimination via LRT ($\alpha = 0.05$) on ML-fit models. Five
interactions are tested and dropped; Region × Nutrients and
Region × Status are retained.

```{r}
#| echo: false
#| include: false

# Backward elimination: drop non-significant interactions via LRT
fit_no_nut_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status + amd * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_full_ml <- update(fit_full, REML = FALSE)
fit_no_nut_status_ml <- update(fit_no_nut_status, REML = FALSE)
lrt1 <- anova(fit_no_nut_status_ml, fit_full_ml)

fit_no_amd_status <- lmer(
  logp1_fruits ~ reg * nutrient * amd +
    reg * status +
    rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_amd_status_ml <- update(fit_no_amd_status, REML = FALSE)
lrt2 <- anova(fit_no_nut_status_ml, fit_no_amd_status_ml)

fit_no_three_way <- lmer(
  logp1_fruits ~ reg * nutrient + nutrient * amd + reg * amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_three_way_ml <- update(fit_no_three_way, REML = FALSE)
lrt3 <- anova(fit_no_three_way_ml, fit_no_amd_status_ml)

fit_no_reg_amd <- lmer(
  logp1_fruits ~ reg * nutrient + nutrient * amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_reg_amd_ml <- update(fit_no_reg_amd, REML = FALSE)
lrt4 <- anova(fit_no_three_way_ml, fit_no_reg_amd_ml)

fit_no_nut_amd <- lmer(
  logp1_fruits ~ reg * nutrient + amd +
    reg * status + rack + (1 | popu) + (1 | gen),
  data = Arabidopsis
)
fit_no_nut_amd_ml <- update(fit_no_nut_amd, REML = FALSE)
lrt5 <- anova(fit_no_reg_amd_ml, fit_no_nut_amd_ml)

fit_reduced <- fit_no_nut_amd
anova_reduced <- anova(fit_reduced, type = 3)

# ---- LRT summary ----
lrts <- list(lrt1, lrt2, lrt3, lrt4, lrt5)
lrt_summary <- data.frame(
  Step = seq_along(lrts),
  Test = c(
    "Nutrients × Status",
    "AMD × Status",
    "Region × Nutrients × AMD",
    "Region × AMD",
    "Nutrients × AMD"
  ),
  Chisq   = round(sapply(lrts, function(x) x$Chisq[2]), 2),
  Df      = sapply(lrts, function(x) x$Df[2]),
  P_Value = signif(sapply(lrts, function(x) x$`Pr(>Chisq)`[2]), 3)
)

# ---- Reduced-model ANOVA ----
anova_tbl <- as.data.frame(anova_reduced)
anova_tbl$Effect <- rownames(anova_tbl)
rownames(anova_tbl) <- NULL
anova_tbl <- anova_tbl %>%
  dplyr::select(Effect, NumDF, DenDF, `F value`, `Pr(>F)`) %>%
  dplyr::mutate(
    DenDF = round(DenDF, 1),
    `F value` = round(`F value`, 2),
    `Pr(>F)` = signif(`Pr(>F)`, 3)
  )
```

\small

::: columns

::: column

**Backward elimination (LRT)**

```{r}
#| echo: false
kable(
  lrt_summary,
  align = "clccc",
  col.names = c("Step", "Interaction dropped", "Chisq", "Df", "p"),
  row.names = FALSE
)
```

:::

::: column

**Reduced model: Type III tests**

```{r}
#| echo: false
kable(
  anova_tbl,
  align = "lcccc",
  col.names = c("Effect", "df1", "df2", "F", "p"),
  row.names = FALSE
)
```

:::

:::

# 3.3 Model Reduction via AIC / BIC

```{r}
#| echo: false
#| include: false
#| cache: true

# Main-effects baseline (all main effects + rack + random effects)
base <- "logp1_fruits ~ reg + nutrient + amd + status + rack + (1 | popu) + (1 | gen)"

treatment_ints <- list(
  "Reg x Nut"       = "reg:nutrient",
  "Reg x AMD"       = "reg:amd",
  "Nut x AMD"       = "nutrient:amd",
  "Reg x Nut x AMD" = "reg:nutrient:amd"
)
nuisance_ints <- list(
  "Reg x Stat" = "reg:status",
  "Nut x Stat" = "nutrient:status",
  "AMD x Stat" = "amd:status"
)

# Hierarchy: if 3-way is in, all three 2-way must also be in.
t_combos <- do.call(expand.grid, replicate(4, c(FALSE, TRUE), simplify = FALSE))
names(t_combos) <- names(treatment_ints)
t_valid <- with(t_combos,
  !`Reg x Nut x AMD` | (`Reg x Nut` & `Reg x AMD` & `Nut x AMD`)
)
t_combos <- t_combos[t_valid, , drop = FALSE]

n_combos <- do.call(expand.grid, replicate(3, c(FALSE, TRUE), simplify = FALSE))
names(n_combos) <- names(nuisance_ints)

models <- list()
for (i in seq_len(nrow(t_combos))) {
  for (j in seq_len(nrow(n_combos))) {
    t_in <- names(treatment_ints)[as.logical(t_combos[i, ])]
    n_in <- names(nuisance_ints)[as.logical(n_combos[j, ])]
    included <- c(t_in, n_in)
    if (length(included) == 0) {
      label <- "Main effects only"
      form <- base
    } else {
      label <- paste(included, collapse = " + ")
      terms <- c(treatment_ints[t_in], nuisance_ints[n_in])
      form <- paste(base, paste(terms, collapse = " + "), sep = " + ")
    }
    models[[label]] <- lmer(as.formula(form), data = Arabidopsis, REML = FALSE)
  }
}

aic_bic <- data.frame(
  Model = names(models),
  AIC   = sapply(models, AIC),
  BIC   = sapply(models, BIC)
)
aic_bic$dAIC <- aic_bic$AIC - min(aic_bic$AIC)
aic_bic$dBIC <- aic_bic$BIC - min(aic_bic$BIC)
aic_bic <- aic_bic[order(aic_bic$AIC), ]
```

::: columns

::: {.column width="42%"}

\small

- Enumerate all subsets of 7 candidate interactions
- **Hierarchy enforced**: if 3-way is in, all three 2-way interactions must also be in
- Main effects always retained; Rack additive
- 72 models (9 treatment × 8 nuisance); fit with ML
- AIC-preferred model adds **Nutrients × AMD** vs. the LRT-reduced model; not significant by LRT
- Retain LRT-reduced model for interpretability

:::

::: {.column width="58%"}

\footnotesize

```{r}
#| echo: false
knitr::kable(
  head(aic_bic, 10), digits = 2, row.names = FALSE,
  caption = "Top 10 of 72 models, ranked by AIC"
)
```

:::

:::

# 3.4 Alternative GLMM with Negative Binomial Family

::: columns

::: {.column width="45%"}

\small

- Total Fruits is a zero-heavy count → fit a **negative binomial GLMM** on the original scale
- Same fixed/random effects as the reduced LMM
- Goal: check whether a count model changes conclusions
- **Result:** similar conclusions for key fixed effects; LMM retained for interpretability

:::

::: {.column width="55%"}

\footnotesize

```{r}
#| echo: false
#| warning: false
#| message: false

fit_nb <- glmmTMB(
  total.fruits ~ reg * nutrient + reg * status + amd + rack +
    (1 | popu) + (1 | gen),
  data = Arabidopsis,
  family = nbinom2
)

nb_anova <- as.data.frame(car::Anova(fit_nb, type = 3))
nb_anova$Effect <- rownames(nb_anova)
rownames(nb_anova) <- NULL

nb_tbl <- nb_anova[, c("Effect", "Chisq", "Df", "Pr(>Chisq)")]
names(nb_tbl) <- c("Effect", "Chisq", "Df", "p-value")
nb_tbl$Chisq <- round(nb_tbl$Chisq, 2)
nb_tbl$`p-value` <- ifelse(
  nb_tbl$`p-value` < 0.0001,
  "<0.0001",
  formatC(nb_tbl$`p-value`, format = "f", digits = 4)
)

kable(nb_tbl, align = "lccc",
      caption = "Type III tests: NB GLMM")
```

:::

:::

# 3.5 Variance Components

Research question 3 asks how much variation in plant response is attributable
to differences between populations and maternal seed families within regions.
We answer it by partitioning the random-effect variance from the reduced LMM
and testing each component via LRT.

::: columns

::: column

```{r}
#| echo: false

vc <- as.data.frame(VarCorr(fit_reduced))
vc <- vc[, c("grp", "vcov")]
vc$pct  <- 100 * vc$vcov / sum(vc$vcov)
vc$vcov <- round(vc$vcov, 3)
vc$pct  <- round(vc$pct, 1)
names(vc) <- c("Component", "Variance", "% of total")
kable(
  vc, row.names = FALSE,
  caption = "Variance components (reduced LMM)"
)
```

:::

::: column

```{r}
#| echo: false

rt <- as.data.frame(lmerTest::ranova(fit_reduced))
rt$Term <- rownames(rt)
rt <- rt[, c("Term", "LRT", "Df", "Pr(>Chisq)")]
rt$LRT <- round(rt$LRT, 2)
rt$`Pr(>Chisq)` <- ifelse(
  rt$`Pr(>Chisq)` < 0.001,
  "<0.001",
  formatC(rt$`Pr(>Chisq)`, format = "f", digits = 3)
)
kable(
  rt, row.names = FALSE,
  col.names = c("Term", "LRT", "Df", "p-value"),
  caption = "LRT for random effects"
)
```

:::

:::

Based on the variance components and LRT results, we attribute approximately 16% of the 
variance in fruit production to differences between populations. The LRT for the 
variance contributed by maternal seed families is not significant, suggesting that this 
level of grouping does not explain a significant portion of the variability in the response.

# 4.1 Assumption Checks

::: columns

::: column

**Residual Plots**

- No strong nonlinear pattern  
- Pattern reflects zero-inflation (not model misspecification)  
- Residuals approximately normal  
- Slight variance increase at higher fitted values  

:::

::: column

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
par(mar = c(2.5, 2.5, 0.5, 0.5))
plot(fit_reduced,
     type = c("p", "smooth"),
     col.line = 1,
     xlab = "Fitted values",
     ylab = "Std. residuals",
     main = "")
```
:::

:::

# 4.2 Variance Checks

- Brown–Forsythe (Levene) test:
  - Some heteroscedasticity across Region and Status
  - Variance ratios < 1.5 → not severe

- Large sample size (n = 625):
  - Test is sensitive to small differences*

**Final Assessment**

- Normality: approximately satisfied  
- Independence: satisfied by design  
- Equal variance: mild violations  

→ Overall, model assumptions are **reasonably satisfied**

# 4.3 Assumption Checks: Q–Q plot

::: columns
::: column

- Q–Q plot shows residuals $≈$ normal  
- Shapiro–Wilk test (p = 0.03) attributed to large sample size (n=625) and zero-inflation, not severe non-normality

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
# Formal normality test on residuals
resids <- resid(fit_reduced, scaled = TRUE)
shapiro_test <- shapiro.test(resids)

shapiro_test
```
:::

::: column

```{r, fig.width=4.5, fig.height=3, out.width='100%', echo=FALSE}
# Q-Q plot
lattice::qqmath(fit_reduced, main = "Normal Q-Q plot")
```

:::

:::

# 4.4 Assumption Checks: Variance

- Scale-location plot:
  - Slight increase in variance at moderate fitted values  
  - No severe heteroscedasticity pattern  

```{r, fig.width=5, fig.height=2.8, out.width='80%', fig.align='center', echo=FALSE}
scale_loc_plot = function(m) {
  plot(m, sqrt(abs(resid(.))) ~ fitted(.), 
       type = c("p", "smooth"),
       xlab = "Fitted",
       ylab = "Sqrt(|Residuals|)",
       main = "")
}
scale_loc_plot(fit_reduced)

```

# 4.5 Conclusion

::: columns

::: column

**Levene Test**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(car)

lev_results <- data.frame(
  Factor = c("Region", "Nutrients", "Status", "AMD"),
  p_value = c(
    leveneTest(resid(fit_reduced) ~ Arabidopsis$reg)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$nutrient)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$status)[1,3],
    leveneTest(resid(fit_reduced) ~ Arabidopsis$amd)[1,3]
  )
)

lev_results$p_value <- signif(lev_results$p_value, 3)
knitr::kable(lev_results, align = "lc")

```

**Final Assessment**

```{r, echo=FALSE}
var_reg <- tapply(resid(fit_reduced), Arabidopsis$reg, var)
var_status <- tapply(resid(fit_reduced), Arabidopsis$status, var)

ratio_reg <- max(var_reg) / min(var_reg)
ratio_status <- max(var_status) / min(var_status)

data.frame(
  Group = c("Region", "Status"),
  Variance_Ratio = round(c(ratio_reg, ratio_status), 2)
) %>% knitr::kable(align = "lc")

```

:::

::: column

**Conclusion**

- Log transformation addresses skewness and zeros 

- Mild heteroscedasticity, but not severe  

- GLMM confirms robustness of results  


→ Model assumptions reasonably satisfied  

:::
:::


# 4.6 Hypothesis Tests — Type III F-tests

```{r}
#| echo: false
anova_reduced <- anova(fit_reduced, type = 3)
tbl <- as.data.frame(anova_reduced)
tbl$Effect <- rownames(tbl)
tbl <- tbl[, c("Effect", "F value", "NumDF", "DenDF", "Pr(>F)")]
tbl$`F value` <- round(tbl$`F value`, 3)
tbl$DenDF    <- round(tbl$DenDF, 2)
tbl$`Pr(>F)` <- ifelse(
  tbl$`Pr(>F)` < 0.001,
  "<0.001",
  formatC(tbl$`Pr(>F)`, format = "f", digits = 3)
)
kable(
  tbl, row.names = FALSE,
  col.names = c("Effect", "F", "df1", "df2", "p-value"),
  align = "lcccc",
  caption = "Type III F-tests for log(Total Fruits + 1)"
)
```

**Significant Effects**

- Significant: **Nutrients, AMD, Rack, and Status**  
- Significant interactions:  
  - Region × Nutrients  
  - Region × Status
  
# 4.7 Post-hoc Analysis (EMMs)

Pairwise contrasts on the log(Total Fruits + 1) scale for significant effects.

::: columns

::: {.column width="48%"}

\footnotesize

**Nutrients within Region**

```{r}
#| echo: false
emm_nutrient <- emmeans(fit_reduced, pairwise ~ nutrient | reg)
kable(
  as.data.frame(emm_nutrient$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Reg", "Contrast", "Est.", "SE", "p")
)
```

\vspace{0.6em}

**AMD (main effect)**

```{r}
#| echo: false
emm_amd <- emmeans(fit_reduced, pairwise ~ amd)
kable(
  as.data.frame(emm_amd$contrasts)[, c("contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Contrast", "Est.", "SE", "p")
)
```

:::

::: {.column width="52%"}

\footnotesize

**Status within Region**

```{r}
#| echo: false
emm_status <- emmeans(fit_reduced, pairwise ~ status | reg)
kable(
  as.data.frame(emm_status$contrasts)[, c("reg", "contrast", "estimate", "SE", "p.value")],
  digits = 3, row.names = FALSE,
  col.names = c("Reg", "Contrast", "Est.", "SE", "p")
)
```

:::

:::

# 4.8 Interaction Plots

**Visualizing Significant Interactions**

The interaction plots below illustrate how the effects of **Nutrients** and **Status** on fruit production vary across **Regions**.

::: columns
::: column
**Region × Nutrients**

```{r, fig.width=4.5, fig.height=3.5, out.width='100%', echo=FALSE}
# Region × Nutrients interaction
Arabidopsis %>%
  group_by(reg, nutrient) %>%
  summarise(
    mean_fruits = mean(logp1_fruits),
    se = sd(logp1_fruits) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = factor(nutrient, labels = c("Low (1 prill)", "High (8 prills)")),
             y = mean_fruits,
             color = reg,
             group = reg)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_fruits - se, ymax = mean_fruits + se),
                width = 0.1, linewidth = 0.8) +
  labs(x = "Nutrient Level",
       y = "Mean log(Total Fruits + 1)",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 10))
```

- All regions benefit from high nutrients
- SP shows the steepest increase
- SW shows the smallest response

:::

::: column

**Region x Status**

```{r, fig.width=4.5, fig.height=3.5, out.width='100%', echo=FALSE}
# Region × Status interaction
Arabidopsis %>%
  group_by(reg, status) %>%
  summarise(
    mean_fruits = mean(logp1_fruits),
    se = sd(logp1_fruits) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = status,
             y = mean_fruits,
             color = reg,
             group = reg)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_fruits - se, ymax = mean_fruits + se),
                width = 0.15, linewidth = 0.8) +
  labs(x = "Plant Status",
       y = "Mean log(Total Fruits + 1)",
       color = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 10),
        axis.text.x = element_text(angle = 15, hjust = 1))
```

- SP: Normal plants outperform Petri.Plate
- SW: Petri.Plate plants have lowest production
- NL: Less differentiation among groups

:::

:::

# 5.1 Summary of Findings

- **Nutrients, AMD, Rack, and Status** were all found to have significant impacts on fruit 
production. Plants grown in the presence of high nutrients and in the absence of AMD produced significantly more fruits than those under low nutrients and exposed to AMD.

- **Region alone was not significant**, but its role became important through interactions 
with other predictors (Nutrients and Status).

- Overall, fruit production in *Arabidopsis* was shaped by both **experimental treatments** 
and **hierarchical biological differences** among regions and populations.

# 5.2 Research Question Conclusions

1. How do the reproductive responses of Arabidopsis thaliana to apical meristem damage (AMD) vary across different regions, nutrient levels, and transplant sources?

    - No significant interactions between AMD and Region, Nutrients, or Status were observed, suggesting the effect of AMD on fruit production was consistent across these factors.

2. Are there significant interactions between region, nutrient levels, and transplant sources in determining a plant's fruit production?

    - Yes, significant interactions were found between Region × Nutrients and Region × Status, indicating that the effects of nutrient availability and transplant source varied across regions.

3. How much of the variability in fruit production can be attributed to differences between populations and maternal seed families within regions?

    - Approximately 16% of the variance in fruit production was attributable to differences between populations, while the variance attributed to maternal seed families was not significant.

# 5.3 Limitations \& Future Work

- Residual diagnostics suggested **some heteroscedasticity across regions and status groups**, although the variance differences were not extreme.

- Because the response variable is a **count with many zeros**, transformation-based Gaussian modeling may not fully capture the data structure.

- While the linear mixed-effects model was interpretable and supported the main conclusions, future work should further investigate **generalized mixed models for count outcomes**.

# 5.4 References

- Banta, J. A., Stevens, M. H. H., & Pigliucci, M. (2010).  
  *A comprehensive test of the “limiting resources” framework applied to plant tolerance to apical meristem damage*.  
  Oikos, 119(2), 359–369.


- Dean, A., Voss, D., & Draguljić, D. (2017).  
  *Design and Analysis of Experiments* (2nd ed.). Springer.


- STAT 642 Course Handouts, including:
  - Handout 6: Random Effects Models
  - Handout 9: Mixed Models and Variance Components
  - Handout 10: Model Construction and Inference
